Modeling and Simulation of Neocortical Micro- and Mesocircuitry. Part II: Physiology and Experimentation

Version 1

Abstract

Summary

Cortical dynamics underlie many cognitive processes and emerge from complex multi-scale interactions, which are challenging to study in vivo. Large-scale, biophysically detailed models offer a tool which can complement laboratory approaches. We present a model comprising eight somatosensory cortex subregions, 4.2 million morphological and electrically-detailed neurons, and 13.2 billion local and mid-range synapses. In silico tools enabled reproduction and extension of complex laboratory experiments under a single parameterization, providing strong validation. The model reproduced millisecond-precise stimulus-responses, stimulus-encoding under targeted optogenetic activation, and selective propagation of stimulus-evoked activity to downstream areas. The model’s direct correspondence with biology generated predictions about how multiscale organization shapes activity; for example, how cortical activity is shaped by high-dimensional connectivity motifs in local and mid-range connectivity, and spatial targeting rules by inhibitory subpopulations. The latter was facilitated using a rewired connectome which included specific targeting rules observed for different inhibitory neuron types in electron microscopy. The model also predicted the role of inhibitory interneuron types and different layers in stimulus encoding. Simulation tools and a large subvolume of the model are made available to enable further community-driven improvement, validation and investigation.

Introduction

Neuroscience aims to characterize the dynamics of the brain and understand how they emerge as a product of anatomy and physiology. This is challenging however, due to the complexity of the brain’s multi-scale organization and its study in situ. Large-scale, data-driven, biophysically-detailed models (Markram et al., 2015; Billeh et al., 2020) offer a tool which can complement laboratory investigation. This “bottom-up” approach combines detailed constituent models into a 3D model of a given brain volume, thus consolidating disparate, multi-scale data sources. Such models aim to provide a general tool for multi-scale investigation; access and manipulation of any model component enables predictions about how these components combine to shape emergent dynamics. With increasing biological data, model refinement and validation must be continuous. This necessitates 1) open sourcing of models with high quality software tools for iterative, community-driven refinement and validation, and 2) rigorous methodology for model parameterization and validation, such that model iterations can be meaningfully compared. Here and in our companion paper (Reimann et al., 2022a) we present our methodology for parameterizing the anatomy and physiology of a cortical model, and validating its emergent dynamics. This enabled predictions about how cortical activity is shaped by high-dimensional connectivity motifs in local and mid-range connectivity, and spatial targeting rules by inhibitory subpopulations. Elsewhere, this has allowed the presented model to be used to study the formation of cell assemblies (Ecker et al., 2024), functional synaptic plasticity (Ecker et al., 2023), the role of non-random connectivity motifs on reliability (Egas Santander et al., 2024), the composition of high-level electrical signals such as the EEG (Tharayil et al., 2024), and how spike sorting biases population codes.

Specifically, we built and validated a model of the entire non-barrel primary somatosensory cortex (nbS1) comprising eight subregions. Whilst our previous data-driven, biophysically detailed model (Markram et al., 2015) provided insights at the scale of a single cortical column (Reimann et al., 2013, 2017, 2022b; Nolte et al., 2019, 2020; Newton et al., 2021), the new model is ∼140 times larger and to our knowledge offers the first simulations of in vivo-like spontaneous and stimulus-evoked activity in a biophysically detailed cortical model with interregion connectivity. In the companion paper, we introduce the anatomical model (Fig. 1, Step 1), describing how neuron morphologies were placed within an atlas-based geometry and connected through local and mid-range synapses. Here, we describe our improved techniques to model and validate the electrical properties of neurons and synapses (Figure 1, Steps 2 & 3 (Reva et al., 2023; Barros-Zulaica et al., 2019)), and to compensate for input from missing brain areas (Figure 1, Steps 4). These improvements enabled enhanced validation of emerging in vivo-like activity (Figure 1, Steps 5), including the reproduction and extension of five published studies in rodent sensory cortex under a single in vivo-like regime (Figure 1, Steps 6).

Overview of the physiology and simulation workflow.

1. Anatomical model: Summary of the anatomical nbS1 model described in the companion paper. 2. Neuron physiology: Neurons were modeled as multi-compartment models with ion channel densities optimised using previously established methods and data from somatic and dendritic recordings of membrane potentials in vitro. 3. Synaptic physiology: Models of synapses were built using previously established methods and data from paired recordings in vitro. 4. Compensation for missing synapses: Excitatory synapses originating from outside nbS1 were compensated with noisy somatic conductance injection, parameterized by a novel algorithm. 5. In vivo-like activity: We calibrated an in silico activity regime compatible with in vivo spontaneous and stimulus-evoked activity. 6. In silico experimentation: Five laboratory experiments were recreated. Two were used for calibration and three of them were extended beyond their original scope. 7. Open Source: Simulation software and a seven column subvolume of the model are available on Zenodo (see data availability statement). Data generalisations: Three data generalisation strategies were employed to obtain the required data. Left: Mouse to rat, middle: Adult to juvenile (P14) rat, right: Hindlimb (S1HL) and barrel field (S1BF) subregions to the whole nbS1. Throughout the figure, the corresponding purple icons show where these strategies were used.

Validating that network activity emerges from the same interactions driving in vivo dynamics, requires 1) an approach to tackle the large parameter space without overfitting and 2) comparison of emerging dynamics with laboratory experiments. With respect to 1), we strongly prefer parameterization with directly measured quantities over fitting parameters to yield the correct emerging activity. Additionally, where fitting is applied, we adhere to a principle of compartmentalization of parameters. That is, once a parameter has been parameterized at one biological level, it is no longer a free parameter at a higher level. For example, the maximal conductance of a synapse is fit to biological amplitudes of postsynaptic potentials, but is then never updated in the process of reaching in-vivo like population activity. If a connection is valid on the single-cell level, its contribution at the population level should be equally valid. As a result, the emerging in-vivo like activity is the consequence of only 10 free parameters representing the extrinsic input into the individual layers from other brain regions. These parameters were fitted using a novel methodology presented here and were in accordance with the mean number of missing synapses for each population (Felleman and Van Essen, 1991; Harris et al., 2019; Gao et al., 2022). With respect to 2), we conducted the following validations: Spontaneous activity reproduced layer-wise in vivo firing rates (or a specified proportion of in vivo firing rates to account for in vivo recording bias (Wohrer et al., 2013)), varied along a spectrum of asynchronous to synchronous activity, exhibited spatially structured fluctuations and produced long-tailed firing rate distributions with sub 1 Hz peaks as in vivo (Wohrer et al., 2013; Buzśaki and Mizuseki, 2014). Under the same parameterization, the model also reproduced precise millisecond dynamics of layer-wise populations in response to simple stimuli and demonstrated selective propagation of stimulus-evoked activity to downstream areas. Again under the same parameterization, we reproduced and extended more complex experiments through accurate modeling of targeted optogenetic stimulation and lesions. Importantly, we highlight where emergent activity shows discrepancies with in vivo activity, to guide future data-driven model refinement.

The model generated a number of predictions (Table S1), including about the role of different layers in driving layer 2/3 stimulus responses and how inhibitory interneuron types encode contrast, synchronous and rate-coded information. Additionally, with access to the full structural connectome and tools for precisely editing it (Pokorny et al., in prep), we were able to make predictions about the relationship between structure and function. For example, we predict that an increase in the prevalence of non-random connectivity motifs towards deeper layers leads to a matched increase in spiking correlations, and that subregions more strongly innervated by mid-range connectivity have higher correlated activity locally. Additionally, we utilized new analysis of inhibitory connectivity (Schneider-Mizell et al., 2023) in the MICrONS electron microspcopy dataset (MICrONSConsortium et al., 2021) to make predictions about the role of specific targeting rules by different inhibitory neuron types. Particularly, we compared activity in the original connectome to a rewired connectome with increased perisomatic targeting by PV+ neurons, and increased targeting of inhibitory populations by VIP+ neurons. Inhibitory populations were more strongly inhibited (increasingly towards central layers) and required more non-local drive to reach the firing rates observed in vivo. Evoked responses increased and decreased in superficial and deeper excitatory populations respectively, suggesting layerspecific roles of the more specific inhibitory targeting.

To provide a framework for further studies and integration of experimental data, a sizable part of the model and simulation tools are made available (Figure 1, Step 7). The detailed modeling approach provides a one-to-one correspondence with most types of experimental data, allowing different datasets to be readily integrated. Due to the incredible speed of discovery in neuroscience an integrative model will always be lagging behind the latest available data. We believe the solution is to provide a scientifically solid, validated model with clearly characterized strengths and weaknesses, along with the tools to advance or customize it for individual projects. We have therefore also made our tools for building and improving the model openly available: www.github.com/BlueBrain.

Results

The companion paper describes the full anatomical nbS1 model containing 4.2M morphological neuron models, connected through 9.1B local synapses and 4.1B mid-range synapses (Reimann et al., 2022a). Each neuron is modeled as a multi-compartmental model belonging to one of 60 morphological types (m-types; Figure 2A1), and is either an instance or statistical variant of an exemplar in a pool of 1,017 morphological reconstructions. Neurons were placed (orientated towards the surface) within an atlas-based geometry (Figure 2A2) based on estimated layer-wise density profiles of different morphological types. Local connectivity is based on axo-dendritic overlap with neighbouring neurons, whilst mid-range connectivity combines data on interregion connectivity and laminar innervation profiles. Thalamocortical afferents were also modeled based on laminar innervation to the barrel cortex. To simulate emergent activity it is necessary to model and validate the electrical properties of these neurons and synapses (Figure 1 Steps 2 & 3). These modeling steps are based on published methods and data sources (Table 1), and are summarized first. The method for compensating for missing synapses and the remaining results are then described (Figure 1 Steps 4-6).

Modeling and validation of neuron physiology.

A1: Example excitatory neuron morphologies. See companion paper for exemplar morphologies of all morphological types. A2: Example morphologies placed within the atlas-based volume. Axons shown in grey. B: Optimized conductance densities for two exemplary e-types. B1: cADpyr e-type on L5 TPC:A m-type, B2: bSTUT and cSTUT e-types on L5 NBC m-type. (The L5 NBC m-type is combined with more e-types than the two shown, see panel D.) Morphologies were visualized with NeuroMorphoVis (Abdellah et al., 2018). Neurite diameters are enlarged (x3) for visibility. Soma and dendrites in black, axon in red. C: e-types used in the model. As used in Markram et al. (2015) and similar to the Petilla terminology (Ascoli et al., 2008)). D: me-type composition. Heatmap shows the proportion of e-types for each m-type. Each me-type is assigned to one of the three I subpopulations. Assignments depending not only on m-type are highlighted by a green box. Particularly, for BTC and DBC m-types, the cACint e-type belongs to the Sst+ subpopulation. E: Validation of dendritic physiology of all L5 TTPCs. Panel reproduced from Reva et al. (2023). E1: Validation of back-propagating action potential (bAP) amplitude (i.e., the dependence of bAP amplitude on distance from the soma) for basal (green) and apical (blue) dendrites. Reference data (in red) comes from Stuart and Sakmann (1994); Larkum et al. (2001) (apical) and Nevian et al. (2007) (basal). Lines show exponential fits for the in silico (green and blue) and in vitro (red) data. Color bar indicates dendritic diameter. E2: Validation of EPSP attenuation. Reference data (in red) comes from Berger et al. (2001) (apical) and Nevian et al. (2007) (basal). Lines and colorbar same as in E1.

Modeling and validation of neuron physiology

Electrical properties of single neurons were modeled by optimising ion channel densities in specific compartment-types (soma, axon initial segment (AIS), basal dendrite, and apical dendrite) (Figure 2B) using an evolutionary algorithm (IBEA; Van Geit et al., 2016) so that each neuron recreates electrical features of its corresponding electrical type (e-type) under multiple standardized protocols. The methodology and resulting electrical models are described in Reva et al. (2023) (see also Methods). Electrical features included firing properties (e.g. spike frequency, interspike interval), action potential waveforms (e.g. fall and rise time, width) and passive properties (e.g. input resistance). The optimisation was performed for a subset of neuron models. The resulting ion channel densities were generalised to other neuron models of the same e-type. For excitatory neurons such generalization was only made to neurons within the same layer. The resulting electrical activity of each neuron was validated against the corresponding electrical features, including the characteristic firing properties of the 11 e-types (Figure 2C). Model neurons with a mean zscore (over the electrical features) further than two standard deviations from the experimental mean were discarded.

For each of the 60 morphological types (m-types), the corresponding fractions of etypes were determined from experimental data as in Markram et al. (2015), resulting in 208 morpho-electrical types (me-types; Figure 2D). Correspondence with predominant expression of biological markers PV (parvalbumin), Sst (somatostatin) or 5HT3aR (serotonin receptor 3A) was determined based on me-type, as previously done in Markram et al. (2015) (Figure 2D). Finally, for layer five thick-tufted pyramidal cell (L5 TTPC) morphologies, we found that dendritic electrical features, namely the attenuation of back propagating action potentials and excitatory postsynaptic potentials (EPSPs), reproduced experimental measurements (Figure 2E, Reva et al., 2023).

The required data to constrain region-specific me-type distributions was not available when building the model; to our knowledge Yao et al. (2023) is the first study that explored the question systematically (in mouse cortex). The atlas-based geometry does however impose heterogeneity in morphologies that e-types are paired ith: only morphologies that are in line with the region-specific local curvature of the region are used throughout the volume. We demonstrate the effect of geometry on morphological composition in the companion paper.

Modeling and validation of synaptic physiology

Synapses are modeled with a stochastic version of the Tsodyks-Markram model (Tsodyks and Markram, 1997; Markram et al., 1998; Fuhrmann et al., 2002; Loebel et al., 2009), upgraded to feature multi-vesicular release as described in Barros-Zulaica et al. (2019) and Ecker et al. (2020). The model assumes a pool of available vesicles that is utilized by a presynaptic action potential, with a release probability dependent on the extracellular calcium concentration ([Ca2+]o; Ohana and Sakmann, 1998; Rozov et al., 2001; Borst, 2010). Additionally, single vesicles spontaneously release with a low frequency as an additional source of variability. The utilization of vesicles leads to a postsynaptic conductance with bi-exponential kinetics. Short-term plasticity (STP) dynamics in response to sustained presynaptic activation are either facilitating (E1/I1), depressing (E2/I2), or pseudo-linear (I3). E synaptic currents consist of both AMPA and NMDA components, whilst I currents consist of a single GABAA component, except for neurogliaform cells, whose synapses also feature a slow GABAB component. The NMDA component of E synaptic currents depends on the state of the Mg2+ block (Jahr and Stevens, 1990), with parameters fit to cortical recordings from Vargas-Caballero and Robinson (2003) by Chindemi et al. (2022).

The workflow for determining a dense parameter set for all synaptic pathways, starting with sparse data from the literature, is described for the use case of hippocampal CA1 in Ecker et al. (2020) and briefly in the Methods. We combined data sources used in Markram et al. (2015), with a large number of recent data sources (Qi and Feldmeyer, 2016; Barros-Zulaica et al., 2019; Yang et al., 2020, 2022). The resulting pathway-specific parameters are listed in Tables S2, S3, S4, the most common short-term dynamics are depicted in Figure 3A1-2, and the assignment of STP profiles to different pathways are shown in Figure 3A3. Postsynaptic potential (PSP) amplitudes and their coefficient of variation (CV; std/mean) closely matched their biological counterparts (r = 0.99, n = 27; Figure 3B1; Table S5 and r = 0.63, n = 10; Figure 3C1; Table S6, respectively). The dense parameter set also allowed prediction of PSP amplitudes and CVs for all cortical pathways (Figure 3B2, C2). The frequency of miniature postsynaptic currents (mPSCs) were also in line with in vitro measurements (r = 0.92, n = 5; Figure 3D; Table S7).

Modeling and validation of synaptic physiology.

Caption on the following page. A1: Exemplary pair of L5 TTPCs (visualized with NeuroMorphoVis (Abdellah et al., 2018)). Presynaptic cell in gray, postsynaptic cell in red, synapses between them in purple. Neurite diameters are enlarged (x3) for visibility and axons were cut to fit into the figure. Pre- and postsynaptic voltage traces on the top right. A2: Exemplary postsynaptic traces with different STP profiles. A3: Assignment of STP profiles to viable pathways. (Pathways were considered viable if there were at least 10 connections in all eight subregions of the model.) B1: Validation of first PSP amplitudes (see also Table S5). Dashed gray line represents perfect correlation between experimental and model values. Error bars show one standard deviation (also for C1, D and F). B2: Predicted PSP amplitudes of all viable pathways in the circuit. Postsynaptic cells were held at -70 mV using an in silico voltage-clamp. Means were calculated over 100 pairs of neurons with 35 repetitions each. C1 and C2: same as B1 and B2, but showing the CV of the first PSP amplitude (corresponding Table is S6). D: Validation of mPSC frequencies (see also Table S7). E: Location of synapses from VPM fibers (purple) and POm fibers (red) on 38 neurons (dark gray) in a 5 µm radius column (visualized with BioExplorer). F: Validation of thalamocortical EPSP amplitudes as in B1. The four pathways used for the validation are marked with a black rectangle on H1 to its right. G: EPSP latencies (time from presynaptic spike to the rise to 5% of peak EPSP amplitude in the postsynaptic trace). H1 Left: mean VPM evoked EPSP amplitudes on postsynaptic cell types (over 50 pairs). Right: Comparison of normalized in silico amplitudes (normalized by L4 E as in Sermet et al., 2019) to in vitro reference data from Sermet et al. (2019). Heatmap shows model minus reference values, thus positive values indicate a higher normalized EPSP amplitude in our model than in the reference experimental dataset. H2: same as H1 but for POm (normalized by L5 E as in Sermet et al., 2019).

The anatomical model includes fibers from the thalamus, based on fibres projecting to the barrel cortex from the ventral posteriomedial (VPM) and posteriormedial (POm) thalamic nuclei. These fibres make synaptic contacts within a radius of the fiber probabilistically based on laminar innervation profiles (see Reimann et al., 2022a; Figure 3E). Compared to the previous model (Markram et al., 2015), POm projections are added and the physiology of synapses from VPM improved (see Methods; Figure 3E, F). Latencies of layer-wise EPSPs increase with distance from the thalamus (Figure 3G). Additionally, thalamocortical EPSP amplitudes normalized relative to a single population were compared to normalized EPSPs in response to optogenetic stimulation targeting bundles of thalamic fibers (Sermet et al., 2019). This provided contrasting insights, however. For example, whilst VPM to L6 I EPSPs match the initial validation data (Figure 3F), VPM to L6 PV+ responses appear too strong relative to other populations (Figure 3H1). The results suggest that the model’s POm to L5 E pathway is too weak, when compared to other POm to E and all POm to PV+ pathways (Figure 3H2, right).

Defining subvolumes and populations

To enable smaller simulations and targeted analyses, we defined standardized partitions comprising all neurons (and their connections) contained in a subvolume of the full nbS1 model. We decomposed the model into full-depth hexagonal prisms of a certain diameter that we call hexagonal subvolumes. These are slightly curved, following the geometry of the cortex, and have an intact layer structure. When the diameter of the hexagons is 520 µm, comparable to the size of the single cortical column model from (Markram et al., 2015), we call these subvolumes columns. Taken together, a central column and the six columns surrounding it define a seven column subvolume (Figure 4A). Additionally, when we separately analyze the E and I neurons in different layers, we refer to them as neuron populations of the model or a subvolume. Groups of I neurons that predominantly express either PV, Sst or 5HT3aR markers, are referred to as inhibitory subpopulations.

Spontaneous activity: calibration, in vivo-like dynamics, linking structure to function.

A: Seven column subvolume. B: OUµ and OUσ of somatic conductance injection (B1) parameterized by population (B2) determine FRs (B3). Target FRs equal to in vivo firing rates multiplied by PFR. Inset: Table summarising meta-parameters. C: Target PFR (x-axis) plotted against the observed PFR (y-axis) for each E (red) and I (blue) population after calibration. The number of sides of the shape indicates the layer (triangle used for L2/3). D: Euclidean distance between target and observed PFR values (over populations) decreases over iterations for two meta-parameter combinations (upper). Values after final iteration shown for all meta-parameter combinations (lower; dashed line shows termination condition). E: (left) Firing rates, (centre) mean conductance injection and (right) OUµ by population for the non-bursting simulations (1 line per simulation; E and I values separated). F: Mean missing number of synapses vs. the mean conductance injection, averaged over all combinations, for each population. Markers as in C. G1: Max normalised histograms for two meta-parameter combinations (5 ms bin size, 1σ Gaussian smoothing). G2: Effect of meta-parameters on correlation between E and I histograms (5 ms bin size, 1σ Gaussian smoothing; central column). G3: Fourier analysis of spontaneous activity for the 59 non-bursting simulations. H: Activity in flat space over seven consecutive 100 ms windows. Activity smoothed (Gaussian kernel, σ = 1pixel). I: Left: Correlation r-value between the histograms of layer-wise E and I populations for the 59 non-bursting simulations. Right: RC/U by population for the 59 non-bursting simulations. J: Top: The mean node participation of E and I populations in each layer for dimension one (left) and six (right) vs. RC/U for the parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4 and PFR = 0.3. Markers as in C. Line shows linear fit. Bottom: Correlation r-values between mean node participation and RC/U when neurons with highest node participation are not included in the calculation of mean node participation (via a sliding threshold).

Compensating for missing brain regions

Initial simulations produced no activity, and it was necessary to compensate for missing excitatory input from neurons external to the model. We estimated that the number of missing synapses for the full nbS1 and seven hexagon subvolume as ∼ two and seven times the number of internal synapses, based on Oh et al. (2014). We model the effect of these missing synaptic inputs explicitly as time-varying and statistically independent somatic conductance injections, using Ornstein-Uhlenbeck (OU) processes that mimic aggregated random background synaptic inputs (Destexhe et al., 2001; Figure 4B1; see also Discussion). The mean (OUµ) and standard deviation (OUσ) are defined as a percentage of an individual cell’s input conductance at its resting membrane potential. The choice of OUµ and OUσ for different populations and [Ca2+]o (Figure 4B2), determine the model’s emergent dynamics and mean firing rates (Figure 4B3). Initial simulations showed that using the same value of OUµ and OUσ for all neurons makes some populations highly active whilst others are silent, or produces network-wide bursts. Finding population-specific OU parameters which produce stable in vivo-like activity is challenging, however, due to the non-linearity and computational cost of simulations.

We developed an algorithm that efficiently calibrates population-specific OU parameters, for different values of [Ca2+]o, which affects the strength and reliability of synaptic transmission (see “Synaptic Physiology”). Additionally, it fixes the ratio OUσ/OUµ, denoted ROU, across all neurons to specific values representing the amount of noise in the extrinsic inputs (Figure 4B2). As extracellularly-derived firing rates are known to be severely overestimated to an unknown degree (Olshausen and Field, 2006; Wohrer et al., 2013; Buzśaki and Mizuseki, 2014), the algorithm finds parameters which produce realistic inter-population firing rate ratios at different global levels of activity (Figure 4B3). Specifically, it targets mean spontaneous firing rates of neuron populations that are a constant fraction PFR of in vivo reference values, for 10 PFR values between 0.1 and 1. We refer to [Ca2+]o, ROU and PFR as meta-parameters. The initial calibration was made for the seven hexagonal subvolume, and later generalised to the full nbS1 model.

The algorithm was first run for [Ca2+]o = 1.1 mM, ROU = 0.4. Target firing rates for the 10 PFR values were reached (Figure 4C) after five iterations of 10 simulations of 6.5 seconds (Figure 4D, left upper). Only three iterations were needed when calibration began using previously calibrated parameters for different ROU and [Ca2+]o (Figure 4D, right upper). This allowed us to parameterize extrinsic input for a range of combinations of the three meta-parameters (Figure 4D, lower & 4E, left) within biologically relevant ranges ([Ca2+]o: 1.05 mM - 1.1 mM (Jones and Keep, 1988; Massimini and Amzica, 2001; Amzica et al., 2002; Gonzalez et al., 2022); PFR: 0.1 - 1.0; ROU : 0.2 - 0.4 (Destexhe et al., 2001)). We found that the resulting mean conductance injections were highly population-specific (Figure 4E, centre). The range of OUµ values between populations was on average 5.2 times higher than the range of OUµ values for a single population across meta-parameter combinations (Figure 4E, right). This was expected, as the missing mid-range innervation is known to have specific and heterogeneous laminar profiles (Felleman and Van Essen, 1991; Harris et al., 2019; Gao et al., 2022). We estimated the number of missing synapses per neuron assuming a total density of E synapses of 1.1 synapses/µm, based on mean spine densities (Larkman, 1991; Datwani et al., 2002; Kawaguchi et al., 2006), and sub-tracting the number of synapses present in the model. We confirmed a strong correlation between this measure and required conductance injection (Figure 4F).

In vivo-like spontaneous activity

Histograms of spontaneous activity for two meta-parameter combinations are shown in Figure 4G1 (and for all combinations in Video 1). The two combinations have the same population firing rates (PFR = 0.9) but the first simulation uses a higher ROU and lower synaptic transmission reliability ([Ca2+]o). In the second simulation, we found larger fluctuations, correlated between E and I populations across layers. Overall, increasing [Ca2+]o and PFR, and decreasing ROU, increases the amplitude of correlated fluctuations (Video 1; Figure 4G2) and increases power at lower frequencies (Figure 4G3). This supports the notion that decreasing ROU or increasing [Ca2+]o shifts the model along a spectrum from externally (noise) driven to internally driven correlated activity.

We confirmed that spontaneous firing rate distributions were long-tailed with peaks below 1 Hz for all of the meta-parameter combinations, in line with in vivo activity (Figure S3). In the horizontal dimensions, under higher correlation regimes, fluctuations are spatially coordinated and global within the central hexagon of the simulated subvolume (Figure 4H; Video 2). In depth, the size of fluctuations and level of correlated activity increases from supragranular to infragranular layers (Figure 4G1, I, left; Video 1), suggesting that activity in deeper layers of the model is more internally driven. We characterized the effect of recurrent connectivity by measuring the layer-specific decrease in firing rates when neurons in the model are disconnected from each other. As expected, we found that the effect (quantified by the ratio between connected and disconnected firing rates: RC/U) increased from suprato infragranular layers (Figure 4I, right).

High-dimensional connectivity motifs predict layer-wise spontaneous activity

By analysing the underlying network structure, we observed a corresponding gradient in the topology of intra-layer synaptic connectivity from supra- to infragranular layers. We calculated the mean node participation of neuron populations in various dimensions (see Methods; Figure 4J, top). For dimension one, this is simply the degree of a neuron; for dimensions above one, this generalizes the notion to counting participation in dense, directed motifs of increasing size (directed simplices; Reimann et al., 2017). We found that correlations of this measure with the ratio of connected and disconnected firing rates increased with dimension (Figure 4J, bottom, 100% of data), indicating the importance of large directed simplices in shaping activity. Curiously, the correlations for higher dimensions were driven by a small number of neurons with a very high node participation. This was indicated by a sharp drop of correlation when neurons above a given value were excluded (Figure 4J, bottom). This demonstrates high variability between neurons, in line with biology, both structurally (Towlson et al., 2013; Nigam et al., 2016) and functionally (Wohrer et al., 2013; Buzśaki and Mizuseki, 2014).

Stimulus-responses reproduce in vivo dynamics with millisecond-scale precision

We compared stimulus-responses with in vivo barrel cortex responses to both single whisker deflections and active whisker touches under anaesthetized (Reyes-Puerta et al., 2015) and awake states (Yu et al., 2019) respectively. While the model is of non-barrel somatosensory regions, this nevertheless provides validations of overall excitability and the laminar structure of responses reflecting general trends of cortical processing. We activated a percentage (FP) of the VPM fibers projecting to the central column of the seven column subvolume (Figure 5A1). For each selected thalamic fiber, spike times were drawn from VPM PSTHs (peristimulus time histogram) recorded in vivo for the two stimulus types (Figure 5A2; Methods). For both stimulus types, a single thalamic stimulus was presented 10 times at 1 s intervals at three intensities (FP : 5%, 10%, 15%).

Layer-wise population responses to single whisker deflection and active whisker touch stimuli closely match in vivo millisecond dynamics and response amplitudes.

A1: Stimuli are simulated by activating a proportion (FP) of fibers projecting to the central hexagon of the seven column subvolumne. A2: Each fiber is assigned a spike time from a PSTH of VPM activity recorded in vivo for either a single whisker deflection or active whisker touch paradigm. B: Spiking activity and histograms (baseline and max normalised) for each layer-wise E and I population for the parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4, PFR = 0.3 and FP = 10% for a 2.5s section of the 10 whisker deflection test protocol. C: Trial-averaged PSTHs (baseline and max normalised) in response to the single whisker deflection stimulus for all of the parameter combinations which passed the initial criteria, and the in vivo references. D: Spatiotemporal evolution of the trial-averaged stimulus response in flat space for the same parameter combination in B. E: Left: Latencies of different layer-wise E and I populations for all criteria passing combinations in response to single whisker deflection stimuli. Right: Latencies for inhibitory interneuron subpopulations in response to active whisker touch stimuli. Correponding in vivo references are shown. F: Heatmap showing the effect of the meta-parameters on the difference of RE for the entire central column from the corresponding in vivo reference. Parameter combinations which failed the criteria tests are masked. G: Left: Ratio RE between maximum evoked response and spontaneous baseline activity by population. One line per simulation. Simulation line colour shows the difference of RE for the entire central column from the corresponding in vivo reference (color values shown in G). Thick black line shows median of in silico values. White shows in vivo reference. Right: Normalised ratios of each simulation by dividing by RE for the entire central column.

For the parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4, PFR = 0.3 and FP = 10%, E and I populations in each layer show clear responses on each trial (Figure 5B; three trials shown), except for L1. Videos 3 and 4 show all simulated combinations of the four meta-parameters. Out of 72 parameter combinations, 21 passed an initial assessment of in vivo similarity to the in vivo data based on latencies and decays (see Methods, Figure 5C, S4). Responses remained localized to within 600 µm of the stimulus location (Figure 5D, Video 5). In vivo, layer-wise I populations precede the corresponding E populations and response latencies increase with distance from thalamus, except for the fast response of L4 I. The latencies and decays are matched by criteria passing simulations (Figure 5C, E, left), except for a longer L4 I latency, and shorter and longer sustained responses of L2/3 E and L4 E. PV+ and Sst+ subpopulations respond reliably in layers 2-6, but the 5HT3aR+ subpopulation responses are weaker and less reliable (Figure S4). Under the active whisker touch paradigm, PV+ latencies precede Sst+ latencies in vivo, and either precede or are simultaneous with Sst+ latencies in silico (Figure 5E, right).

The passing parameter combinations form a contiguous region in the parameter space (Figure 5F), characterized by combinations of low PFR, [Ca2+]o and FP with higher ROU, which displayed weakly correlated spontaneous activity (compare Figure 5F with Figure 4G2). This offers a prediction that true in vivo firing rates are within the range of these lower FRs. The remaining results apply to the criteria passing simulations. We characterized the magnitude of responses by RE, the ratio between the peak of the evoked response and the pre-stimulus baseline activity level. Assuming that extracellular sampling biases are similar during spontaneous and evoked activity, RE provides a bias-free metric for comparing response magnitude with in vivo. Decreasing ROU and PFR, and increasing [Ca2+]o and FP led to a global increase of RE, and results matched in vivo closely for a large part of the parameter space (Figure 5F).

Interestingly, the parameter combination closest to in vivo was central in the parameter space of criteria passing simulations (Figure 5G, left). Values of RE, normalized by the mean RE across populations, were remarkably constant (Figure 5G, right), indicating that variation of the parameters provides a global scaling of response magnitude. Moreover, I populations matched the in vivo reference closely, while the responses of L2/3 E were slightly greater, and for L4 E and L5 E slightly weaker. The region of the parameter space where responses match in vivo for the active whisker touch stimulus was the same as for single whisker deflections (Figure S5). Latencies were also similar to in vivo, with the main discrepancy being slower responses in L4 (E and I). With respect to the metaparameters, we found that the best match for the anesthetized state was characterized by lower firing rates (PFR) and reduced external noise levels (lower ROU) compared to the awake active whisker touch paradigm (white areas in Figures 5F and S5B). These results fit expectation for anaesthetized and awake states respectively, and provide predictions of their spontaneous firing rates, noise levels and corresponding in silico regimes.

Response sparsity (i.e. the proportion of neurons spiking at least once following a stimulus) varied across populations and meta-parameter combinations, but was around the range of 10-20% reported in vivo in the barrel cortex (Barth and Poulet (2012); Figure S4D). Interestingly, excitatory population responses were sparser than the corresponding inhibitory layer-wise populations across layers. The observed heterogeneous layer-wise population sparsities offer a prediction for in vivo. Moreover, the majority of responding cells spike only once to single whisker whisker deflections (Figure S4E), as reported in vivo (Isbister et al., 2021).

Validation and prediction through reproduction and extension of complex laboratory experiments using a single model parameterization

Building upon the in vivo-like activity states, we combined various simulation techniques to recreate laboratory experiments from the literature (see Methods). For these experiments, we use the parameter combination: [Ca2+]o = 1.05 mM, ROU = 0.4 and PFR = 0.3, as this combination supports realistic in vivo-like responses to the two stimulus types over a range of fiber percentages (FP).

Exploring the canonical model: layer-wise contributions to L2/3 stimulus-responses

In the canonical model of the cortex (reviewed e.g., in Lübke and Feldmeyer, 2007; Feldmeyer, 2012) information from the thalamus arrives to L4, then propagates to L2/3, from there to L5 (which serves as the main cortico-cortical output layer) and lastly to L6 (which projects back to the thalamus). The coordinated action of all layers of S1 is required for high-level behavioral tasks (Park et al., 2020). As the canonical model is based on the highest layer-wise density of axons, it cannot describe all interactions in the cortex. For example VPM does not only innervate L4, but also the border of L5 and L6, as well as the bottom of L2/3 (Meyer et al., 2010; Constantinople and Bruno, 2013; Sermet et al., 2019).

To study how L4 contributes to the stimulus preference of L2/3 PCs, Varani et al. (2022) used optogenetic inactivation of L4 PCs during whisker stimulation and quantified the changes in the subthreshold response of L2/3 PCs. They found, that the early phase of the subthreshold response significantly differed from the control condition, if the whisker was deflected in either the most or the least preferred direction (see the top and bottom rows of their Figure 5B, C). From this they concluded that both L4 and VPM contribute to the direction tuning of L2/3 PCs. After reproducing their experimental conditions in silico (Figure 6A,B; Figure S6; see Methods) we confirmed that we can reproduce their results, i.e., subthreshold responses of L2/3 PCs decreased, when L4 PCs were inhibited (Figure 6C for preferred direction whisker stimulation; see Methods).

Reproducing and extending detailed experiments exploring the canonical model and encoding of contrast, rate-coded and synchronous information.

First experiment: Reproducing the effect of optogenetic inactivation of L4 pyramidal cells on L2/3 stimulus responses and predicting the role of other pathways through circuit lesions. A: Schematics of whisker kinetics and VPM fiber rates during 500 ms long whisker hold stimulus. Fraction of VPM fibers coding for each kinetic feature are taken from Petersen et al. (2008). B: Mimicking the effect of activation of the Halo inhibitory opsin in silico. Injected somatic hyperpolarizing current mimicking opsin activation (top), and the resulting somatic voltage trace from a combination of injected conductance, current, and synaptic PSPs from the network (bottom). C: Comparison of average traces from selected L2/3 PCs (see Methods) in control (black) and optogenetically inhibited (green) conditions. D: Same as C, but instead of mimicking the optogenetic inhibition of L4 PCs, only the connections to L2/3 PCs are “cut” (compare inset with the one in C). The right part depicts connections systematically cut from PCs in all layers, while the left shows L4 only for a better visual comparison with the conditions of Varani et al. (2022) in C. Second experiment: Contrast tuning modulation by optogenetic activation of PV+ interneurons, and encoding of synchronous and rate-coded by interneuron subtypes. E: Spatial distribution of firing rates of VPM fibers projecting to a single simulated column at different points in time, corresponding to linear drifting gratings with a temporal frequency of ftemp = 2 Hz and a spatial frequency of fspat = 0.001 cycles/µm. F: Firing rate signal of a single VPM fiber corresponding to a random sequence of drifting grating patterns with different contrast levels C, as indicated by the legend. All optogenetic stimuli targeting PV+ (or Sst+) interneurons were completely overlapping with the grating stimulus intervals, as indicated by the blue bars. G1: Spiking activity and PSTH of an exemplary PV+ interneuron over 10 repetitions (trials) in response to a grating stimulus at maximum contrast (C = 1.0) without (control) and with medium strength (150%) optogenetic stimululation. H1: Contrast tuning modulation of an exemplary PV+ interneuron by different levels of PV+ optogenetic stimulation, ranging from 0% (control) to 300%. The individual markers denote actual data points (mean SD over 10 repetitions, normalized to baseline response at maximum contrast) while curves illustrate sigmoidal fits to these data points. G2/H2: Same as C1/D1, but for exemplary PCs. H2 inset: Illustration of sigmoidal parameters m, n, Rmax, and c50.

Third experiment: Encoding of synchronous and rate-coded by interneuron subtypes. I: A binary signal is encoded either through changes in the rate or synchronicity of optogenetic pulses applied to a set of PCs. We define a direct correspondence between single optogenetic pulses and single spikes. Spiking activity is measured in PV+, Sst+ and 5HT3aR+ cells, and the mutual information between the binned activity of individual cells and the original binary signal is measured. J: Visualization of the rate (upper) and synchronous (lower) coded stimulus experiments stimulating 1000 PCs, showing the binary signal (top), input neuron spike trains for 40 neurons (middle), and responses of the three L2/3 neuron types (bottom). K: Upper: Results for the rate coded stimulus experiment. Left: Mutual information between spiking activity and the binary signal (one point for each cell that spiked at least once). Only activity in the 50 ms following the change of the binary signal is considered. Cells with mutual information not significantly different from that of a shuffled control are coloured grey. Centre: Same as left but considering all time bins. Right: The effect of the number of stimulus neurons on the mean single cell mutual information for each subpopulation. Lower: The same as upper but for the synchronous stimulus type.

We then leveraged our in silico setup to study what Varani et al. (2022) could not, because of methodological limitations. In our reading, the authors aimed to test how direct E connections from L4 PCs to L2/3 PCs influence the stimulus representation in L2/3. This connection can not specifically be blocked in vivo, instead (95% of) the L4 PC population is inhibited (as well as some lower L3 PCs). In our setup we could selectively block the connection and found almost the same result (compare Figure 6C and D, left). This extends the conclusion of Varani et al. (2022): L4 PCs contribute to the stimulus preference of L2/3 PCs via direct E connections, and not via disynaptic inhibition.

The authors also discussed studying L5 PCs’ contribution to L2/3 responses (as a large fraction of L5 PC axons terminate in L2/3), but this is infeasible with current mouse lines. Leveraging our model, we found that L5 contributes much less to subthreshold L2/3 traces than L4 (Figure 6D, right). Extending to other presynaptic layers, we found that the contribution of L2/3 is similar to that of L4, whereas inputs from L6 are negligible (Figure 6D, right). Whilst mouse lines targeting L5 PCs might arrive soon (which could validate our prediction), blocking L2/3 connections between L2/3 cells without hyperpolarizing the same L2/3 population seems only achievable in silico.

Contrast tuning by inhibitory interneuron sub-types under optogenetic activation

Shapiro et al. (2022) compared the modulatory effects of optogenetic activation of interneuron subtypes on contrast tuning of neurons in V1. Whilst visual regions differ from somatosensory regions, the study provided an opportunity to explore the generality of interactions between E and I subpopulations. We therefore created a version of the study in silico, presenting spatio-temporally modulated patterns with various contrast levels through the VPM inputs (Figure 6E,F; see Methods).

Overall, we found within the single simulated column 228 PCs and 259 PV+ interneurons with robust contrast tuning of their responses (see Methods). Within spike detection range of a vertically penetrating extracellular electrode (50 µm; Henze et al. (2000); Buzśaki (2004)), we found 15 PCs and 13 PV+ interneurons with robust tuning, which is roughly comparable to the original study (average of 8.2 tuned PCs and 3.0 PV+ neurons per mouse Shapiro et al., 2022). Unlike the original study we found no Sst+ neurons with robust tuning.

Additionally, we found the following results, all in line with (Shapiro et al., 2022): The firing rates of both PCs and PV+ neurons were affected by optogenetic activation of the PV+ subpopulation (Figure 6G), resulting in changes to the contrast tuning curves of these populations (Figure 6H). The main effect of the optogenetic manipulation was an increase in contrast tuning for PV+ and a decrease for PCs. The tuning curves could be accurately fit by sigmoidal functions (rPV2 = 0.995 ± 0.008; rPC2 = 0.990 ± 0.012), and consequently the effect of optogenetic activation could be quantified in terms of changes to the fitted parameters (Figure 6H1 inset, S7; Methods). We found great variability between neurons with increasing trends for parameter m in PV+ and c50 in PCs, and decreasing trends for Rmax in PV+ and both m and Rmax in PCs. Unlike in (Shapiro et al., 2022), we found in addition an increasing trend of c50 also in PV+ interneurons (Figure S7).

Comparing three abstract mathematical models (Shapiro et al., 2022; Methods), we found that the saturation additive model best captured the effect of the optogenetic manipulation for PV+ neurons (Figure S10A, B1), as in the reference study. The relative contributions of saturation and addition varied strongly between individual neurons (Figure S10C1). At low intensities of the optogenetic manipulation, this distribution matched qualitatively the results of Shapiro et al. (2022), but at higher intensities we observed a shift towards stronger saturation. Combining the saturation additive model with a simplified, population-level model of PC firing (Shapiro et al., 2022; Methods) provided a good description of the observations (Figure S10B2) with relative contributions of saturation and addition that qualitatively match the reference (Figure S10C2).

Encoding of synchronous and rate-coded signals by inhibitory interneuron sub-types

Prince et al. (2021) studied how different I subpopulations represent signals that are encoded as changes in the firing rate or synchronicity of PCs (Figure 6I). We repeated their study in silico (Figure 6J), but using in vivo-like conditions as the authors have suggested in their Discussion. The original study pointed out the importance of membrane properties and short-term dynamics of synaptic inputs onto the subpopulations in shaping the results. Note, that the lower synaptic reliability in vivo weakens synaptic depression (Borst, 2010), and the ongoing background input leads to higher membrane conductance (Destexhe et al., 2001); therefore our results are expected to differ substantially, and provide an independent prediction for a different dynamic state. While in the original in vitro study stimulation of 10 PCs was sufficient to find a decodable signal, we expected a larger number to be required under our in vivo-like conditions, due to the lower [Ca2+]o, and the higher level of noise (see Methods).

Indeed, we found little discernible mutual information in I spiking for the activation of 10 PCs, but that mutual information increased with the number of activated neurons (Figure 6K, right). For 1000 activated neurons each of the three I subpopulations in L2/3 showed clear but qualitatively unique modulation for both encoding schemes (Figure 6J). As in (Prince et al., 2021) we found differences in the encoding capabilities of the I subpopulations (Figure 6K). We found strongest encoding in PV+ neurons for both schemes, and overall low encoding strength in Sst+ neurons. Further, we found overall low amounts of mutual information for the synchronous stimulus, which is in line with Prince et al. (2021), as they linked encoding of that stimulus to depressing synaptic dynamics, which would be weakened in vivo.

Exploring the effect of inhibitory targeting trends observed in electron microscopy

We next explored how certain trends in inhibitory targeting observed in a recent mouse electron microscopic connectome (Schneider-Mizell et al., 2023; Figure 7A) affect spontaneous and evoked activity. Consequently, an alternative connectome which recreates these trends has been built (Reimann et al., 2022a). We refer to this as the Schneider-Mizell compatible connectome (SM-connectome). It ensures that PV+ neurons target the perisomatic region, VIP+ neurons target other inhibitory neurons, and cells in L1 connect predominantly monosynaptically. Consequently, inhibition onto E neurons is sparser and closer to the soma (Figure 7A2). After validating the presence of these trends, we characterized how this affects the dynamics of the spontaneous and evoked states (Figure 7A3). Note, that individual synapses were recalibrated to preserve the pathway specific reference PSP amplitudes. As expected, the largest change was for PV+ neurons where the average number of vesicles in the release-ready pool NRRP was increased by 2.4 times.

The effect of inhibitory targeting trends observed in electron microscopy on spontaneous and stimulus-evoked activity.

A1: Panel illustrating interneuron synaptic targeting features and caption taken from Schneider-Mizell et al. (2023) (under CC-BY 4.0 license). i) For determining inhibitory cell subclasses, connectivity properties were used such as an axon (green) making synapses (green dots) the perisomatic region of a target pyramidal cell (purple). ii) Dendritic compartment definitions for excitatory neurons. iii) Cartoon of a multisynaptic connection (left) and the synapses within the multisynaptic connection considered ”clumped” along the presynaptic axon (right). A2: Exemplar neurons before and after rewiring showing synapse locations from different inhibitory m-type groupings: Distal targeting (DistTC: martinotti, bipolar, double bouquet cells), Perisomatic targeting (PeriTC: large basket, nest basket cells), Sparse targeting (SparTC: descending axon, neurogliaform, horitzontal axon cells), Inhibitory targeting (InhTC: bitufted, small basket cells). A3: Heatmaps showing the mean fraction of efferent synapses from the inhibitory m-type groupings onto inhibitory neurons, somas, proximal dendrites, apical dendrites, distal dendrites, or that are part of clumped or multisynaptic connections. Heatmaps show mean fractions (left to right) for the original, Schneider-Mizell et al. (2023) characterization of the MICrONS dataset ((MICrONS-Consortium et al., 2021)), SM-connectome, and the change between the original and SM-connectome. B1: Proportional change in mean conductance injection between original and SM-connectome by population. B2: Distribution of ratios between connected and disconnected firing rates (RC/U) by population over non-bursting meta-parameter combinations for original and SM-connectome. B3: Comparison of correlation (r-values) between E and I histograms (5 ms bin size, 1σ Gaussian smoothing; central column) between original and SM-connectome (each point represents a specific meta-parameter combination). C: Scatter plot showing the change in the mean number of afferent inhibitory synapses onto each E (red) and I (blue) population after calibration versus the proportional change in mean conductance injection. The number of sides of the shape indicates the layer (triangle used for L2/3). D1: Proportional change in the the ratio between the peak of the evoked response and the pre-stimulus baseline activity level (RE) between the original and SM-connectome. Plot shows distribution over meta-parameter combinations which passed the the initial assessment of in vivo similarity to the in vivo data based on latencies and decays (see Methods). D2: Change in the latencies of populations peak responses between original and SM-connectome (distribution over same meta-parameter combinations shown in D1).

Inspite of this, the rewiring affected the activity sufficiently that recalibration of missing synaptic compensation was needed to attain the target firing rates for the original meta-parameter combinations (Figure S11). Mean conductances onto I neurons had to be increased, indicating the efficacy of the newly introduced I-I targeting by VIP+ neurons (Figure 7B1). On the other hand, conductances for E neurons remained roughly the same, except for L6. Curiously, we found that the firing rates of I populations were reduced when the internal connectivity was activated compared to when it was inactive, further highlighting the efficacy of the I-I targeting (Figure 7B2, S11). For E neurons in L2-5, activating internal connectivity increased firing rates, but less than in the original connectome. As the number of inhibitory synapses innervating them decreased (Figure 7A2), this indicates that the placement of PV+ synapses closer to the soma increased the efficacy of inhibitory as expected. Taken together, this indicates that in the SM-connectome, I populations are more externally driven and more internally inhibited (as a function of proximity to L4), and L6 E more internally driven during spontaneous activity. Over the different meta-parameter combinations, correlation in the network decreased relative to the original circuit with some exceptions (Figure 7B3, S11). This indicates, that activity in the SM-connectome is shifted slightly towards a more asynchronous state.

By correlating the structural and functional changes we concluded that the increased conductance injection into the I populations can be explained by the increased inhibitory synapse count (Figure 7C). Conversely, the mostly decreased synapse counts for E populations did not lead to a decrease of conductance injection, further indicating that the more perisomatic positioning of PV+ synapses compensates for the reduced synapse counts. The lack of effect onto most of the E populations disappeared in the evoked state (Figure 7D). Overall, responses increase strongly in L23 and slightly decreased in L5 and L6 (Figure 7D1), while latencies of responses increased in every layer (Figure 7D2), suggesting different roles of the more specific inhibitory targeting in different layers.

Mid-range connectivity determines independent functional units and selective propagation of stimulus-responses at larger spatial scales

After calibrating the model of extrinsic synaptic input for the seven column subvolume, we tested to what degree the calibration generalizes to the entire nbS1. Notably, this included addition of mid-range connectivity (Reimann et al., 2022a). The total number of local and mid-range synapses in the model was 9138 billion and 4075 billion, i.e., on average full model simulations increased the number of intrinsic synapses onto a neuron by 45%. Particularly, the final ϕ1.05,0.4 calibrated for the seven column subvolume was used in conjunction with PFR [0.1, 0.15, …, 0.3]. Each of the full nbS1 simulations (using [Ca2+]o = 1.05 mM and ROU = 0.4) produced stable non-bursting activity, except for the simulation for PFR = 0.3, which produced network-wide bursting activity.

Activity levels in the simulations of spontaneous activity were heterogeneous, ranging from a mean firing rate of 0.15 Hz to 0.3 Hz (when activity was averaged in flat space for PFR = 0.15; Figure 8A1; Video 6). The subregion with consistently higher firing rates was the upper-limb representation (S1ULp), separated from subregions of lower activity by the dysgranular zones (S1DZ, S1DZO). This demonstrates that the larger spatial scale of the model supports several independently acting functional units. Locations with higher firing rates were accompanied by higher spiking correlations of the local population (Figure 8A2), indicating activity that was more driven by intrinsic excitation. Unlike in the simulations of smaller volumes, the correlation emerged mostly from short, transient packets of activity in L6 (Figure 8B). Consequently, deviation from the target PFR in those cases was largest in L6 (Figure 8C). Taken together, it appears as if the additional synapses from mid-range connectivity moved the model along the spectrum from extrinsically vs. intrinsically driven, but further for some regions than for others. We confirmed a divergence into two populations: one with correlation slightly increased from the simulations of smaller volumes, and one with almost maximal correlations (Figure 8D). This distinction was determined mostly in L6 with increases of correlations in other layers remaining small.

Full nbS1 simulations: independent functional units, millisecond-precise stimulus-responses, selective propagation of stimulus-evoked activity through mid-range connectivity, linking structure to function.

Caption on following page. A1: Mean FRs across the model. Columns 0 and 39 (subsequently used) are highlighted. A2: Spiking correlations of E/I populations in 240 hexagonal subvolumes (diameter: 460 µm). B: Spiking activity and max normalised histograms for layer-wise E and I populations for the parameter combination in A, and the 0th (upper) and 39th (lower). C: Target vs. observed PFR values for the four non-bursting simulations. The number of sides of each marker corresponds to the layer (triangle represents L2/3). Red: E, blue: I.. D: Correlations between E and I populations by layer for the four non-bursting simulations for the two hexagons (top and bottom). Color of line from light to dark represents PFR. Middle: Distributions of spiking correlations of layer-specific populations in columns (520 µm) of the nbS1 model (black dots; mean values: grey). Compared to the central column when the seven column subvolume is simulated in isolation (green). E: Trial-averaged activity in flatspace following single whisker deflections. Lower: Subset of time windows shown with only the top 2% of bins for each time window after baseline activity level subtracted. F: Trial-averaged PSTHs (baseline and max normalised) for column 0. For FP = 15% and FP = 25% respectively, the evoked responses passed and failed the latency and decay-based criteria tests for similarity to in vivo. Response for FP = 20% passed the tests for all populations except L5 E which showed a more sustained response. G: Trial-averaged response of column 39. H: Correlations of mid-range inputs. Upper: Distribution of spiking correlations of inputs into 240 subvolumes. Calculated at reduced spatial resolution, based on connection counts and correlations between the subvolumes (see Methods). Subvolumes are sorted by the E/I correlation of neurons within them. Lower: Mean of the correlations of inputs into 240 subvolumes vs internal spiking correlation for all subvolumes during spontaneous activity. Black line: linear fit. Calculated based on connection counts and correlations between 50 µm hexagonal subvolumes (see Methods). Data for evoked activity is shown in Figure S12. I: Upper: Spiking correlations of subvolumes against their distances in spontaneous (red) and evoked (orange: Fp = 0.15%, blue: Fp = 0.2%, green: Fp = 0.25) activity. For increased spatial resolution, smaller (58 µm) subvolumes were used, hence correlation values are not comparable to values in B. Lower: Distribution of soma distances of neuron pairs connected by local (orange) or mid-range (blue) connectivity. Green: distances of all pairs, independent of connectivity. J: Upper: Classification of subvolumes based on low or high internal correlation and membership in mid-range rich club Data for evoked activity shown in Figure S12. Lower: E/I correlation of subvolumes for members of the rich club (red) and non-members (blue). K: FRs of subvolumes following a stimulus with FP = 0.15% (top) and FP = 0.2% (bottom). Mean (lines) and SEM (shaded area) over 10 repetitions shown for subvolumes directly innervated by the VPM stimulus (black), strongly innervated by directly innervated subvolumes (with over 106 synapses, red), or by medium strength (over 105 synapses, yellow) or weak (over 104 synapses) indirect innervation. Dashed lines indicate locations of peaks

We repeated the previous single whisker deflection evoked activity experiment in the full model, providing a synchronous thalamic input into the forelimb sub-region (S1FL; Figure 8E; Video 7). Responses in S1FL were remarkably similar to the ones in the seven column subvolume, including the delays and decays of activity (Figure 8F). However, in addition to a localized primary response in S1FL within 350 µm of the stimulus, we found several secondary responses at distal locations (Figure 8E; Video 7), which was suggestive of selective propagation of the stimulus-evoked signal to downstream areas efferently connected by mid-range connectivity. The response of the main activated downstream area (visible in Figure 8E) was confined to L6 (Figure 8G).

As the model is homogeneous across the different subregions in terms of neuronal morphologies and physiology, we expected the differences in spontaneous and evoked activity across the model to be explained by the topology of synaptic connectivity (Reimann et al., 2022a). While the extrinsic input from the OU injections was uncorrelated by design, synaptic input from local connectivity would feature at least some degree of correlation. This led to asking, how correlated was the input from mid-range connections? We estimated for each pair of mid-range connections innervating a 460 µm hexagonal subvolume its spiking correlation and compared the resulting distribution to the correlation within the subvolume itself (see Methods; Figure 8H, upper). We found low correlations for most pairs, but with a significant number of high-correlation outliers for subvolumes with high internal correlations. Moreover, when we considered the mean of the measure, we found that it explains most of the variability in internal correlations, for both spontaneous (Figure 8H, lower) and evoked states (Figure S12).

This led to the following tentative explanation: Local connectivity provides inputs from within 500 µm, which is virtually guaranteed to be correlated to some degree (Figure 8I upper vs. lower, orange). This is a result of connections being locally dense, as they are concentrated within a small neighborhood (Figure 8 lower, green vs. orange). mid-range connectivity originates from further away, sampling from different functional units that may or may not be correlated, and being more diluted (Figure 8I lower, green vs. blue). Consequently, mid-range inputs are likely to act not much different from the uncorrelated OU process, unless this is overcome by a non-random topology of mid-range connectivity.

Specifically, we considered the mid-range connectivity at meso-scale in terms of the graph of synapse counts between 460 µm subvolumes. When the graph was thresholded at eight million synapses, most subvolumes lacked connections at that strength completely, but others formed a single connected component, indicating the formation of a rich club (Colizza et al., 2006; Figure 8J, upper, red and orange). We found that subvolumes with high internal correlation were part of the rich club, or directly adjacent to one of its members (Figure 8J). In the presence of strong thalamic inputs, this could be overcome, as the increased activity level spread along sufficiently strong mid-range connections (Figure 8K), albeit with substantial delays between 60 and 100 ms. This could induce high correlations in strongly connected parts of the model, outside the rich club as well (Figure S12) and shows that the model supports selective propagation of activity to areas efferently connected with mid-range connectivity.

Discussion

We presented novel methods to build, simulate and validate models of cortical tissue that correspond directly with 3D digitised brain atlases. As demonstrated, this enables laboratory experiments to have a simulatable in silico counterpart, and vice versa: predictions made by the model automatically have a precise correspondence in biology. By recreating and extending five laboratory experiments under a single model parameterization, we provided strong model validation and demonstrated the model’s natural versatility. To our knowledge the simulations of the full nbS1 represent the first simulations of stable in vivo-like spontaneous and stimulus-evoked activity in a large-scale biophysically detailed model of multiple cortical subregions connected through local and mid-range connectivity. The model generated an initial set of predictions about the relationship between cortical structure and function, and provides a basis for future studies and predictions. Table S1 describes each quantitative prediction made in this manuscript, and how it might be tested in vivo. We made the model and simulation tools available for use, enabling community driven testing, exploration and refinement.

Excitation from non-modeled brain regions is modeled by injecting somatic conductances and is calibrated for specific populations. For the purpose of calibration, we presented a novel technique, which allows rapid and methodical characterisation and validation of emergent model activity. The efficiency of the technique allows rapid recalibration after changes to the model such as adding new brain regions or changing anatomical details. Connected divided by unconnected firing rates measured during the process are also a prediction of the degree to which the activity of a population is determined by incoming connections from local vs extrinsic sources. We found an increase of the importance of local connectivity from superficial to deeper layers, in line with the canonical view of the cortex (Lübke and Feldmeyer, 2007; Feldmeyer, 2012), which places layers 5 and 6 at the end of local information flow.

Although abstract relative to the model, the injection amplitudes could be predicted from the number of missing synapses, suggesting it provides a functional representation of their anatomical counterparts. The modular model of extrinsic connectivity can be refined or replaced in the future, or when larger fractions of the brain are modeled. Our approach to obtaining in vivo-like activity contrasts with a recent approach in a largescale biophysical model of primary visual cortex of similar scale to our smaller subvolume (Billeh et al., 2020). There, extrinsic inputs were delivered through dendritic synapses instead of somatic injection, which is more anatomically realistic, and intrinsic recurrent weights were adjusted to match extracellularly recorded firing rates. They find a single activity state, instead of the in vivo-compatible spectrum here that allowed us to contrast anesthetized and awake states. The model in general differs from hybrid models, which jointly use biologically-detailed models and point neuron models (Billeh et al., 2020; DuraBernal et al., 2023), and also from the work of Egger et al. (2020) who modeled activation of L5 pyramidal cells by constraining patterns of synaptic input based on receptive fields and predicted anatomical innervation.

Given the complexity of the model, how can we be confident that the its activity is in vivo-like? First, we demonstrated correlated spontaneous dynamics in the form of global and local fluctuations and dynamic E-I balance, as found experimentally (Renart et al., 2010). Second, population firing rate distributions were long-tailed with sub 1 Hz peaks, and were similar for spontaneous and stimulus-evoked activity, as reported (Wohrer et al., 2013). Moreover, mean firing rates below 1 Hz are required by metabolic constraints (Attwell and Laughlin, 2001; Lennie, 2003). In response to simple whisker stimuli, response sparsity, spike counts, and the temporal profile and amplitudes of layer-wise populations were similar to in vivo under anesthetized and awake conditions, for different regions of the meta-parameter space. The effect of the spontaneous meta-parameters on evoked responses, also shows that the networks spontaneous state affects stimulus-responses, as observed in vivo (Isbister et al., 2021). Although from the barrel system, these stimuli offer some of the simplest stimulus-response paradigms, enabling principles of neural dynamics and information processing to be studied; in particular, the correspondence between atomic units of sensory information and neural representations. The model also predicts the number of thalamic fibers stimulated for whisker-flick stimuli. Such validations are crucial for complex models, as they provide context for more complex validations. Moreover, in a rich and complex nonlinear system, the source of discrepancies with in vivo activity are simpler to ascertain under simpler protocols.

For more complex validations, we reproduced protocols from three laboratory experiments. These experiments used a wide array of techniques, which could be accurately recreated due to the model’s close correspondence with biological tissue. Additionally, we could go beyond the original experiments. For example, we predicted an increased number of neurons would be required to have a decodable effect on postsynaptic activity under in vivo conditions over the in vitro conditions of the original experiment. This can be explained by lower synaptic reliability under in vivo conditions. This demonstrates that the approach of modeling entire brain regions can be used to further probe the topics of the original articles and cortical processing. This could be continued in the future using paradigms that take advantage of the multiple regions offered by the model to study processing in cortical hierarchies.

Responses to whisker flick stimuli were closer to biology for values of the PFR metaparameter between 0.1 and 0.5, according with the known inflation of mean firing rates stemming from the bias of extracellular spike sorting techniques towards larger and more active neurons (Olshausen and Field, 2006). Supporting our estimates, patch-clamp experiments show ubiquitous neuronal silence (Crochet and Petersen, 2006), and spontaneous and stimulus-evoked barrel cortex firing rates as much as 10 times lower than for extracellular recordings (0.05-0.15 Hz vs. 0.8-1.5 Hz; Olshausen and Field, 2006). Each recording technique comes with challenges and biases (Shoham et al., 2006; Olshausen and Field, 2006; Barth and Poulet, 2012; Wohrer et al., 2013). Calcium imaging and patchclamp experiments are less biased towards frequently spiking neurons (Wohrer et al., 2013), yet the former can only infer spiking activity and may be biased towards neurons that favor marker expression. Patch-clamp techniques can reduce firing rates, be biased towards larger or more active neurons (Olshausen and Field, 2006); and are limited to recording a few neurons simultaneously, preventing characterizations of single trial population dynamics. An earlier version of the model has been used to simulate cortical extracellular potentials (Reimann et al., 2013). We can now simulate modern extracellular electrodes and compare multi-unit activity and spike sorting results to the in vivo-like ground truth to gain further insights into the potential over-estimation of mean firing rates.

During simulations of the entire nbS1, we found emerging spatial inhomogeneities in network activity. Specifically, we observed sharp transitions in firing rates and pairwise correlations at disgranular zones. This highlights that the model supports several, independently acting functional units. Analysis of spatial correlations suggests a radius of approximately 400 µm for a single unit. This property was not present in the previous intra-regional scale model (Markram et al., 2015), but is rather an emergent property of the interregional scale, demonstrating that we achieved our goal of building a model that can be used to study mid-range interactions between regions. Analyzing the source of the inhomogeneities suggested a prominent role of a non-random rich-club structure of mid-range connections at the meso-scale. Within functional units, we found that inhomogeneous increases of spiking correlations were confined to spontaneous packets of L6 activity.

It is important to highlight where the model deviates from biological data. First, the model omits several known anatomical elements, such as glia and the presence of gap junctions between certain neuron-types, but the model’s spatial context provides a natural scaffold for them. Second, we model only a single electrical type for pyramidal cells, while some literature sources propose several (Gouwens et al., 2019). However, we are not aware of any method which can recreate the electrical features of different types of pyramidal cells using only generic ion channel models. To achieve the firing pattern behavior of more specific electrical types, usually ion channel kinetics are tweaked, and this would validate the compartmentalization of parameters. Our modeling techniques assume generality of rules and parameters, unless indicated by the data. Therefore, violations indicate that biological rules and data may be more specific. For example, model thalamic inputs innervate any dendrite placed in the targeted layers equally, but this did not reproduce the timing of inhibitory subpopulation responses. Stronger innervation of PV+ neurons than other inhibitory neurons may rectify this. However, it is unclear whether innervation should be anatomically or physiologically stronger, as the best available data mixes structural and physiological strength (Sermet et al., 2019).

Finally, when required, we generalized data from different animals and brain regions to build our rat nbS1 model. This is the accepted state-of-the-art in computational neuroscience, e.g., all 19 data-driven models of rodent microcircuitry listed in Figure 2 of the recent review of Ramaswamy (2024) conduct some sort of mixing, including the advanced mouse V1 model of the Allen Institute (Billeh et al., 2020). While, a first truly single species model would be a great advance, it is not required to be the basis of in silico research. Our model (and previous ones) can be used to study the many cortical mechanisms that are common to the closely related organisms rat and mouse. This is illustrated by the follow-up research using the model in the fields of spiking codes (Ecker et al., 2024; Egas Santander et al., 2024), plasticity (Ecker et al., 2023) and electrical signals (Tharayil et al., 2024).

However, we cannot always predict the effect caused by mixing data sources. An interesting example of this is described in the companion paper (Reimann et al., 2022a), stating that certain rules of inhibitory connectivity found in a mouse electron-microscopic (EM) connectome (Schneider-Mizell et al., 2023) cannot be explained by an non-selective pruning axo-dendritic appositions. We demonstrated the utility of the model to explore the effect of any connectivity rule, using the example of a mouse EM-compatible “SMconnectome”. After rewiring, and refitting synaptic, and conductance injection parameters, we found that the purely perisomatic targeting by PV+ neurons and the increased inhibitory targeting by VIP+ neurons shifted how much individual layers were driven by intrinsic and extrinsic excitation. This could not be explored with current techniques in vivo and highlights a strength of in silico experimentation. On the other hand, this data source contradicts some of our rat nbS1 parameters, e.g., the number of synapses per connection from light microscopy (Markram et al., 2015) do not match with the ones obtained from mouse V1 EM. Thus we used the rat data sources for building our baseline model, but make both models available in a simulatable format (see Data Availability).

Supporting information

eLife Assessment

This valuable manuscript reports a large-scale, data-driven, biophysically detailed model of the non-barrel primary somatosensory cortex and generates numerous predictions that can further our understanding of how the multiscale organization of the cortex shapes neural activity. While the approach is solid, many of the findings are obtained using a much smaller portion of the model, which, together with the broad scope of the work, makes the narrative somewhat confusing and the strength of findings not entirely clear.

Reviewer #1 (Public review):

This paper presents a model of the whole somatosensory non-barrel cortex of the rat, with 4.2 million morphologically and electrically detailed neurons, with many aspects of the model constrained by a variety of data. The paper focuses on simulation experiments, testing a range of observations. These experiments are aimed at understanding how the multiscale organization of the cortical network shapes neural activity.

Strengths:

(1) The model is very large and detailed. With 4.2 million neurons and 13.2 billion synapses, as well as the level of biophysical realism employed, it is a highly comprehensive computational representation of the cortical network.

(2) Large scope of work - the authors cover a variety of properties of the network structure and activity in this paper, from dendritic and synaptic physiology to multi-area neural activity.

(3) Direct comparisons with experiments, shown throughout the paper, are laudable.

(4) The authors make a number of observations, like describing how high-dimensional connectivity motifs shape patterns of neural activity, which can be useful for thinking about the relations between the structure and the function of the cortical network.

(5) Sharing the simulation tools and a "large subvolume of the model" is appreciated.

Weaknesses:

(1) A substantial part of this paper - the first few figures - focuses on single-cell and single-synapse properties, with high similarity to what was shown in Markram et al., 2015. Details may differ, but overall it is quite similar.

(2) Although the paper is about the model of the whole non-barrel somatosensory cortex, out of all figures, only one deals with simulations of the whole non-barrel somatosensory cortex. Most figures focus on simulations that involve one or a few "microcolumns". Again, it is rather similar to what was done by Markram et al., 2015 and constitutes relatively incremental progress.

(3) With a model like this, one has an opportunity to investigate computations and interactions across an extensive cortical network in an in vivo-like context. However, the simulations presented are not addressing realistic specific situations corresponding to animals performing a task or perceiving a relevant somatosensory stimulus. This makes the insights into the roles of cell types or connectivity architecture less interesting, as they are presented for relatively abstract situations. It is hard to see their relationship to important questions that the community would be excited about - theoretical concepts like predictive coding, biophysical mechanisms like dendritic nonlinearities, or circuit properties like feedforward, lateral, and feedback processing across interacting cortical areas. In other words, what do we learn from this work conceptually, especially, about the whole non-barrel somatosensory cortex?

(4) Most comparisons with in vivo-like activity are done using experimental data for whisker deflection (plus some from the visual stimulation in V1). But this model is for the non-barrel somatosensory cortex, so exactly the part of the cortex that has less to do with whiskers (or vision). Is it not possible to find any in vivo neural activity data from the non-barrel cortex?

(5) The authors almost do not show raw spike rasters or firing rates. I am sure most readers would want to decide for themselves whether the model makes sense, and for that, the first thing to do is to look at raster plots and distributions of firing rates. Instead, the authors show comparisons with in vivo data using highly processed, normalized metrics.

(6) While the authors claim that their model with one set of parameters reproduces many experimentally established metrics, that is not entirely what one finds. Instead, they provide different levels of overall stimulation to their model (adjusting the target "P_FR" parameter, with values from 0 to 1, and other parameters), and that influences results. If I get this right (the figures could really be improved with better organization and labeling), simulations with P_FR closer to 1 provide more realistic firing rate levels for a few different cases, however, P_FR of 0.3 and possibly above tends to cause highly synchronized activity - what the authors call bursting, but which also could be called epileptic-like activity in the network.

(7) The authors mention that the model is available online, but the "Resource availability" section does not describe that in substantial detail. As they mention in the Abstract, it is only a subvolume that is available. That might be fine, but more detail in appropriate parts of the paper would be useful.

Reviewer #2 (Public review):

Summary:

This paper is a companion to Reminann et al. (2022), presenting a large-scale, data-driven, biophysically detailed model of the non-barrel primary somatosensory cortex (nbS1). To achieve this unprecedented scale of a bottom-up model, approximately 140 times larger than the previous model (Markram et al., 2015), they developed new methods to account for inputs from missing brain areas, among other improvements. Isbister et al. focus on detailing these methodological advancements and describing the model's ability to reproduce in vivo-like spontaneous, stimulus-evoked, and optogenetically modified activity.

Strengths:

The model generated a series of predictions that are currently impossible in vivo, as summarized in Table S1. Additionally, the tools used in this study are made available online, fostering community-based exploration. Together with the companion paper, this study makes significant contributions by detailing the model's constraints, validations, and potential caveats, which are likely to serve as a basis for advancing further research in this area.

Weaknesses:

That said, I have several suggestions to improve clarity and strengthen the validation of the model's in vivo relevance.

Major:

(1) For the stimulus-response simulations, the authors should also reference, analyze, and compare data from O'Connor et al. (2010; https://pubmed.ncbi.nlm.nih.gov/20869600/) and Yu et al .(2016; https://pubmed.ncbi.nlm.nih.gov/27749825/) in addition to Yu et al. 2019, which is the only data source the authors consider for an awake response. The authors mentioned bias in spike rate measurements, but O'Connor et al. used cell-attached recordings, which do not suffer from activity-based selection bias (in addition, they also performed Ca2+ imaging of L2/3). This was done in the exact same task as Yu et al., 2019, and they recorded from over 100 neurons across layers. Combining this data with Yu et al., 2019 would provide a comprehensive view of activity across layers and inhibitory cell types. Additionally, Yu et al. (2016) recorded VPM neurons in the same task, alongside whole-cell recordings in L4, showing that L4 PV neurons filter movement-related signals encoded in thalamocortical inputs during active touch. This dataset is more suitable for extracting VPM activity, as it was collected under the same behavior and from the same species (Unlike Diamond et al., 1992, which used anesthetized rats). Furthermore, this filtering is an interesting computation performed by the network the authors modeled. The validation would be significantly strengthened and more biologically interesting if the authors could also reproduce the filtering properties, membrane potential dynamics, and variability in the encoding of touch across neurons, not just the latency (which is likely largely determined by the distance and number of synapses).

(2) The authors mention that in the model, the response of the main activated downstream area was confined to L6. Is this consistent with in vivo observations? Additionally, is there any in vivo characterization of the distance dependence of spiking correlation to validate Figure 8I?

(3) Across the figures, activity is averaged across neurons within layers and E or I cell types, with a limited description of single-cell type and single-cell responses. Were there any predictions regarding the responses of particular cell types that significantly differ from others in the same layer? Such predictions could be valuable for future investigations and could showcase the advantages of a data-driven, biophysically detailed model.

(4) 2.4: Are there caveats to assuming the OU process as a model for missing inputs? Inputs to the cortex are usually correlated and low-dimensional (i.e., communication subspace between cortical regions), but the OU process assumes independent conductance injection. Can (weakly) correlated inputs give rise to different activity regimes in the model? Can you add a discussion on this?

(5) 2.6: The network structure is well characterized in the companion paper, where the authors report that correlations in higher dimensions were driven by a small number of neurons with high participation ratios. It would be interesting to identify which cell types exhibit high node participation in high-dimensional simplices and examine the spiking activity of cells within these motifs. This could generate testable predictions and inform theoretical cell-type-specific point neuron models for excitatory/inhibitory balanced networks and cortical processing.

Minor:

(1) Since the previous model was published in 2015, the neuroscience field has seen significant advancements in single-cell and single-nucleus sequencing, leading to the clustering of transcriptomic cell types in the entire mouse brain. For instance, the Allen Institute has identified ~10 distinct glutamatergic cell types in layer 5, which exceeds the number incorporated into the current model. Could you discuss 1) the relationship between the modeled me-types and these transcriptomic cell types, and 2) how future models will evolve to integrate this new information? If there are gaps in knowledge in order to incorporate some transcriptome cell types into your model, it would be helpful to highlight them so that efforts can be directed toward addressing these areas.

(2) For the optogenetic manipulation, it would be interesting if the model could reproduce the paradoxical effects (for example, Mahrach et al. reported paradoxical effects caused by PV manipulation in S1; https://pubmed.ncbi.nlm.nih.gov/31951197/). This seems a more relevant and non-trivial network phenomenon than the V1 manipulation the authors attempted to replicate.

Author response:

The following is the authors’ response to the previous reviews.

Public Reviews:

Reviewer #1 (Public review):

This paper presents a model of the whole somatosensory non-barrel cortex of the rat, with 4.2 million morphologically and electrically detailed neurons, with many aspects of the model constrained by a variety of data. The paper focuses on simulation experiments, testing a range of observations. These experiments are aimed at understanding how the multiscale organization of the cortical network shapes neural activity.

Strengths:

(1) The model is very large and detailed. With 4.2 million neurons and 13.2 billion synapses, as well as the level of biophysical realism employed, it is a highly comprehensive computational representation of the cortical network.

(2) Large scope of work - the authors cover a variety of properties of the network structure and activity in this paper, from dendritic and synaptic physiology to multi-area neural activity.

(3) Direct comparisons with experiments, shown throughout the paper, are laudable.

(4) The authors make a number of observations, like describing how high-dimensional connectivity motifs shape patterns of neural activity, which can be useful for thinking about the relations between the structure and the function of the cortical network.

(5) Sharing the simulation tools and a "large subvolume of the model" is appreciated.

We thank the reviewer for these comments and are pleased they appreciated these aspects of the work.

Weaknesses:

(1) A substantial part of this paper - the first few figures - focuses on single-cell and single-synapse properties, with high similarity to what was shown in Markram et al., 2015. Details may differ, but overall it is quite similar.

We thank the reviewer for this useful comment and agree that it is important to better highlight the incremental improvements to the model’s low-level physiology. The validity of any model can continuously be improved at all spatial scales and the validity of emergent network activity increases with improved validity at lower levels. For this reason, we felt it was valuable to improve the low-level physiology of the model.

Regarding neuron physiology, we have added the following in Section 2.1 on page 5:

“2.1 Improved modeling and validation of neuron physiology

Similarly to Markram et al. (2015), electrical properties of single neurons were modelled by optimizing ion channel densities in specific compartment-types (soma, axon initial segment (AIS), basal dendrite, and apical dendrite) (Figure 2B) using an evolutionary algorithm (IBEA; Van Geit et al., 2016) so that each neuron recreates electrical features of its corresponding electrical type (e-type) under multiple standardized protocols. Compared to Markram et al. (2015), electrical models were optimized and validated using 1) additional in vitro data, features and protocols, 2) ion channel and electrophysiological data corrected for the liquid junction potential, and 3) stochastic channels (StochKv3) now including inactivation profiles. The methodology and resulting electrical models are described in Reva et al. (2023) (see Methods), and generated quantitatively more accurate electrical activity, including improved attenuation of excitatory postsynaptic potentials (EPSPs) and back-propagating action potentials.”

And page 8:

“The new neuron models saw a 5-fold improvement in generalizability compared to Markram et al. (2015) (Reva et al., 2023).”

We have also made the descriptions of the improvements to synaptic physiology more explicit in Section 2.2 on page 9:

“2.2 Improved modeling and validation of synaptic physiology

The biological realism of synaptic physiology was improved relative to Markram et al. (2015) using additional data sources and by extending the stochastic version of the Tsodyks-Markram model (Tsodyks and Markram, 1997; Markram et al., 1998; Fuhrmann et al., 2002; Loebel et al., 2009) to feature multi-vesicular release, which in turn improved the accuracy of the coefficient of variations (CV; std/mean) of postsynaptic potentials (PSPs) as described in Barros-Zulaica et al. (2019) and Ecker et al. (2020). The model assumes a pool of available vesicles that is utilized by a presynaptic action potential, with a release probability dependent on the extracellular calcium concentration ([Ca2+]o; Ohana and Sakmann, 1998; Rozov et al., 2001; Borst, 2010). Additionally, single vesicles spontaneously release as an additional source of variability with a low frequency (with improved calibration relative to Markram et al. (2015)). The utilization of vesicles leads to a postsynaptic conductance with bi-exponential kinetics. Short-term plasticity (STP) dynamics in response to sustained presynaptic activation are either facilitating (E1/I1), depressing (E2/I2), or pseudo-linear (I3). E synaptic currents consist of both AMPA and NMDA components, whilst I currents consist of a single GABAA component, except for neurogliaform cells, whose synapses also feature a slow GABAB component. The NMDA component of E synaptic currents depends on the state of the Mg2+ block (Jahr and Stevens, 1990), with the improved fitting of parameters to cortical recordings from Vargas-Caballero and Robinson (2003) by Chindemi et al. (2022).”

(2) Although the paper is about the model of the whole non-barrel somatosensory cortex, out of all figures, only one deals with simulations of the whole non-barrel somatosensory cortex. Most figures focus on simulations that involve one or a few "microcolumns". Again, it is rather similar to what was done by Markram et al., 2015 and constitutes relatively incremental progress.

We thank the reviewer for this comment and have added the following text to the Discussion on page 33 to explain our rationale:

“In keeping with the philosophy of compartmentalization of parameters and continuous model refinement (see Introduction), it was essential to improve validity at the columnar scale (relative to Markram et al. (2015)) as part of demonstrating validity of the full nbS1. Indeed, improved parametrization and validation at smaller scales was essential to parameterizing background input which generated robust nbS1 activity within realistic [Ca2+]o and firing rate ranges. We view this as a major achievement, as it was unknown whether the model would achieve a stable and meaningful regime at the start of our investigation. Whilst we would have liked to go further, our primary goal was to publish a well characterized model as an open resource that others could use to undertake further in-depth studies. In this regard, we are pleased that the parametrization of the nbS1 model has already been used to study EEG signals (Tharayil et al., 2024), as well as propagation of activity between two subregions (Bolaños-Puchet and Reimann, 2024).”

We also make it clearer in the Introduction on page 4 that the improved validation of the emergent columnar regime was essential to stable activity at the larger scale:

“These initial validations demonstrated that the model was in a more accurate regime compared to Markram et al. (2015) – an essential step before testing more complex or larger-scale validations. For example, under the same parameterization we then observed selective propagation of stimulus-evoked activity to downstream areas, and…”

(3) With a model like this, one has an opportunity to investigate computations and interactions across an extensive cortical network in an in vivo-like context. However, the simulations presented are not addressing realistic specific situations corresponding to animals performing a task or perceiving a relevant somatosensory stimulus. This makes the insights into the roles of cell types or connectivity architecture less interesting, as they are presented for relatively abstract situations. It is hard to see their relationship to important questions that the community would be excited about - theoretical concepts like predictive coding, biophysical mechanisms like dendritic nonlinearities, or circuit properties like feedforward, lateral, and feedback processing across interacting cortical areas. In other words, what do we learn from this work conceptually, especially, about the whole non-barrel somatosensory cortex?

We thank the reviewer for this comment and agree that it would be very interesting to explore such topics. In the Introduction on page 4, we have updated the list of papers which have so far used the model for more in depth studies:

“…propagation of activity between cortical areas (Bolaños-Puchet and Reimann, 2024) the role of non-random connectivity motifs on network activity (Pokorny et al., 2024) and reliability (Egas Santander et al., 2024), the composition of high-level electrical signals such as the EEG (Tharayil et al., 2024), and how spike sorting biases population codes (Laquitaine et al., 2024).”

In the Discussion on page 33 we also add our additional thoughts on this topic:

“Whilst we would have liked to go further, our primary goal was to publish a well characterized model as an open resource that others could use to undertake further in-depth studies. In this regard, we are pleased that the parametrization of the nbS1 model has already been used to study EEG signals (Tharayil et al., 2024), as well as propagation of activity between two subregions (Bolaños-Puchet and Reimann, 2024). Investigation, improvement and validation must be continued at all spatial scales in follow up papers with detailed description, figures and analysis, which cannot be covered in this manuscript. Each new study increases the scope and validity of future investigations. In this way, this model and paper act as a stepping stone towards more complex questions of interest to the community such as perception, task performance, predictive coding and dendritic processing. This was similar for Markram et al. (2015) where the initial paper was followed by more detailed studies. Unlike the Markram et al. (2015) model, the new model can also be exploited by the community and has already been used in a number of follow up papers studying (Ecker et al., 2024a,b; Bolaños-Puchet and Reimann, 2024; Pokorny et al., 2024; Egas Santander et al., 2024; Tharayil et al., 2024; Laquitaine et al., 2024). We believe that the number of use cases for such a general model is vast, and is made larger by the increased size of the model.”

(4) Most comparisons with in vivo-like activity are done using experimental data for whisker deflection (plus some from the visual stimulation in V1). But this model is for the non-barrel somatosensory cortex, so exactly the part of the cortex that has less to do with whiskers (or vision). Is it not possible to find any in vivo neural activity data from the non-barrel cortex?

We agree with the reviewer that this is a weakness. We have expanded our discussion of the need to mix data sources to also consider our view for network level activity:

“This paper and its companion paper serve to present a methodology for modeling micro- and mesoscale anatomy and physiology, which can be applied for other cortical regions and species. With the rapid increase in openly available data, efforts are already in progress to build models of mouse brain regions with reduced reliance on data mixing thanks to much larger quantities of available atlas-based data. This also includes data for the validation of emergent network level activity. Here we chose to compare network-level activity to data mostly from the barrel cortex, as well as a single study from primary visual cortex. Whilst a lot of the data used to build the model was from the barrel cortex, the barrel cortex also represents a very well characterized model of cortical processing for simple and controlled sensory stimuli. The initial comparison of population-wise responses in response to accurate thalamic input for single whisker deflections was essential to demonstrating that the model was closer to in vivo, and we were unaware of similar data for nonbarrel somatosensory regions. Moreover, our optogenetic & lesion study demonstrated the capacity to compare and extend studies of canonical cortical processing in the whisker system.”

(5) The authors almost do not show raw spike rasters or firing rates. I am sure most readers would want to decide for themselves whether the model makes sense, and for that, the first thing to do is to look at raster plots and distributions of firing rates. Instead, the authors show comparisons with in vivo data using highly processed, normalized metrics.

We thank the reviewer for this comment and agree that better visualizations of the network activity under different conditions is essential for helping the reader assess the work. In addition to raster plots in Video 1, Video 3, Fig 6, Fig 5C, Fig S9a, S16a, we have additionally:

a) Changed the histograms of spontaneous activity in Fig 4G on page 13 to raster plots for the seven column subvolume for two contrasting meta-parameter regimes.

b) Added 4 new videos (Video 6a,b and 8a,b) showing all spontaneous and evoked meta-parameter combinations in hex0 and hex39 of the nbS1:

We have added improved plots showing the distributions of firing rates in the seven column subvolume on page 74:

With more detailed consideration in the Results on page 15:

“Long-tailed population firing rate distributions with means ∼ 1Hz

To study the firing rate distributions of different subpopulations and m-types, we ran 50s simulations for the meta-parameter combinations: [Ca2+]o: 1.05mM, ROU: 0.4,PFR: 0.3, 0.7 (Figure S4). Different subpopulations showed different sparsity levels (proportion of neurons spiking at least once) ranging from 6.6 to 42.5%. Wohrer et al. (2013) considered in detail the biases and challenges in obtaining ground truth firing rate distributions in vivo, and discuss the wide heterogeneity of reports in different modalities using different recording techniques. They conclude that most evidence points towards longtailed distributions with peaks just below 1Hz. We confirmed that spontaneous firing rate distributions were long-tailed (approximately lognormally distributed) with means on the order of 1Hz for most subpopulations. Importantly the layer-wise means were just below 1Hz in all layers for the PFR = 0.3 meta-parameter combination. Moreover, our recent work applying spike sorting to extracellular activity using this meta-parameter combination found spike sorted firing rate distributions to be lognormally distributed and very similar to in vivo distributions obtained using the same probe geometry and spike sorter (Laquitaine et al., 2024).

(6) While the authors claim that their model with one set of parameters reproduces many experimentally established metrics, that is not entirely what one finds. Instead, they provide different levels of overall stimulation to their model (adjusting the target "P_FR" parameter, with values from 0 to 1, and other parameters), and that influences results. If I get this right (the figures could really be improved with better organization and labeling), simulations withPFR closer to 1 provide more realistic firing rate levels for a few different cases, however, PFR of 0.3 and possibly above tends to cause highly synchronized activity - what the authors call bursting, but which also could be called epileptic-like activity in the network.

We thank the reviewer for this comment. We can now see that the motivation for PFR parameter was introduced very briefly in the results and that the results of the calibration and analysis of the spontaneous activity regime are not interpreted in relation to this parameter.

To address this, we have given more detail where it is first introduced in the Results on page 12:

“to account for uncertainty in the firing rate bias during spontaneous activity from extracellular spike sorted recordings…”

We then reconsider that it represents an unknown bias when interpreting the calibration and spontaneous activity results on page 15:

“We reemphasize that the [Ca2+]o, ROU and PFR meta-parameters account for uncertainty of in vivo extracellular calcium concentration, the nature of inputs from other brain regions and the bias of extracellularly recorded firing rates. Whilst estimates for [Ca2+]o are between 1.0 - 1.1mM (Jones and Keep, 1988; Massimini and Amzica, 2001; Amzica et al., 2002; Gonzalez et al., 2022) and estimates for PFR are in the range of 0.1 - 0.3 (Olshausen and Field, 2006), combinations of these parameters supporting in vivo-like stimulus responses in later sections will offer a prediction for the true values of these parameters. Both these later results and our recent analysis of spike sorting bias using this model (Laquitaine et al., 2024) predict a spike sorting bias corresponding to PFR ∼ 0.3, confirming the prediction of Olshausen and Field (2006).”

And in relation to the stimulus evoked responses on page 17:

“Specifically, simulations with PFR from 0.1 to 0.5 robustly support realistic stimulus responses, with the middle of this range (0.3) corresponding with estimates of in vivo recording bias; both the previous estimates of Olshausen and Field (2006) and from a spike sorting study using this model (Laquitaine et al., 2024).”

Following these considerations, the remainder of the experiments using the seven column subvolume only use a single meta-parameter on page 19.

For the full nbS1 we further discuss the importance of a P_FR value between 0.1 and 0.3 in the Results on page 26:

“Stable spontaneous activity only emerges in nbS1 at predicted in vivo firing rates

After calibrating the model of extrinsic synaptic input for the seven column subvolume, we tested to what degree the calibration generalizes to the entire nbS1. Notably, this included the addition of mid-range connectivity (Reimann et al., 2024). The total number of local and mid-range synapses in the model was 9138 billion and 4075 billion, i.e., on average full model simulations increased the number of intrinsic synapses onto a neuron by 45%. Particularly, we ran simulations for PFR</i ∈ [0.1, 0.15, ..., 0.3] using the OU parameters calibrated for the seven column subvolume for [Ca2+]o = 1.05mM and ROU = 0.4. Each of these full nbS1 simulations produced stable non-bursting activity (Figure 8A), except for the simulation for PFR</i = 0.3, which produced network-wide bursting activity (Video 6). Activity levels in the simulations of spontaneous activity were heterogeneous (Figure 8B, Video 7). In some areas, firing rates were equal to the target PFR, whilst in others they increased above the target (Figure 8C). In the more active regions, mean firing rates (averaged over layers) were on the order of 30-35% of the in vivo references for the maximum non-bursting PFR simulation (target PFR : 0.25). This range of firing rates again fits with the estimate of firing rate bias from our paper studying spike sorting bias (Laquitaine et al., 2024) and the meta-parameter range supporting realistic stimulus responses in the seven column subvolume. This also predicts that the nbS1 cannot sustain higher firing rates without entering a bursting regime.

Finally, we also added to our discussion of biases in extracellular firing rates in the Discussion on page 32:

“This is also inline with our recent work using the model, which estimated a spike sorting bias corresponding to PFR = 0.3 using virtual extracellular electrodes (Laquitaine et al., 2024).”

We also thank the reviewer for pointing out that we did not define the term “bursting” in the main text. We have added the following definition and discussion in the Results on page 15:

“Note that the most correlated meta-parameter combination [Ca2+]o: 1.1mM, ROU: 0.2, PFR: 1.0 produced network-wide “bursting” activity, which we define as highly synchronous all or nothing events (Video 1). Such activity, which may be characteristic of epileptic activity, can be studied with the model but is not the focus of this study.”

(7) The authors mention that the model is available online, but the "Resource availability" section does not describe that in substantial detail. As they mention in the Abstract, it is only a subvolume that is available. That might be fine, but more detail in appropriate parts of the paper would be useful.

Firstly, we are pleased to say that the full nbS1 model is now available to download, in addition to the seven hexagon subvolume. In the manuscript, we have:

a) Added to the Introduction at the bottom of page 4:

“To provide a framework for further studies and integration of experimental data, the full model is made available with simulation tools, as well as a smaller subvolume with the optional new connectome capturing inhibitory targeting rules from electron microscopy”.

b) Updated the open source panel of Figure 1:

Secondly, we thank the reviewer for noticing that the description of the available model is not well described in the “Resource availability” statement and have addressed this by:

a) Adding the following to the “Resource availability” statement on page 36:

“Both the full nbS1 model and smaller seven hexagon subvolume are available on Harvard Dataverse and Zenodo respectively in SONATA format (Dai et al., 2020) with simulation code. DOIs are listed under the heading ``Final simulatable models'' in the Key resources table. An additional link is provided to the SM-Connectome with instructions on how to use it with the seven hexagon subvolume model.”

b) Creating a new subheading in the “Key resources table” titled: “Final simulatable models” to make it clearer which links refer to the final models.

Reviewer #2 (Public review):

Summary:

This paper is a companion to Reimann et al. (2022), presenting a large-scale, data-driven, biophysically detailed model of the non-barrel primary somatosensory cortex (nbS1). To achieve this unprecedented scale of a bottom-up model, approximately 140 times larger than the previous model (Markram et al., 2015), they developed new methods to account for inputs from missing brain areas, among other improvements. Isbister et al. focus on detailing these methodological advancements and describing the model's ability to reproduce in vivo-like spontaneous, stimulus-evoked, and optogenetically modified activity.

Strengths:

The model generated a series of predictions that are currently impossible in vivo, as summarized in Table S1. Additionally, the tools used in this study are made available online, fostering community-based exploration. Together with the companion paper, this study makes significant contributions by detailing the model's constraints, validations, and potential caveats, which are likely to serve as a basis for advancing further research in this area.

We thank the reviewer for these comments, and are pleased they appreciate these aspects of the work.

Weaknesses:

That said, I have several suggestions to improve clarity and strengthen the validation of the model's in vivo relevance.

Major:

(1) For the stimulus-response simulations, the authors should also reference, analyze, and compare data from O'Connor et al. (2010; https://pubmed.ncbi.nlm.nih.gov/20869600/) and Yu et al .(2016; https://pubmed.ncbi.nlm.nih.gov/27749825/) in addition to Yu et al. 2019, which is the only data source the authors consider for an awake response. The authors mentioned bias in spike rate measurements, but O'Connor et al. used cell-attached recordings, which do not suffer from activity-based selection bias (in addition, they also performed Ca2+ imaging of L2/3). This was done in the exact same task as Yu et al., 2019, and they recorded from over 100 neurons across layers. Combining this data with Yu et al., 2019 would provide a comprehensive view of activity across layers and inhibitory cell types. Additionally, Yu et al. (2016) recorded VPM neurons in the same task, alongside whole-cell recordings in L4, showing that L4 PV neurons filter movement-related signals encoded in thalamocortical inputs during active touch. This dataset is more suitable for extracting VPM activity, as it was collected under the same behavior and from the same species (Unlike Diamond et al., 1992, which used anesthetized rats). Furthermore, this filtering is an interesting computation performed by the network the authors modeled. The validation would be significantly strengthened and more biologically interesting if the authors could also reproduce the filtering properties, membrane potential dynamics, and variability in the encoding of touch across neurons, not just the latency (which is likely largely determined by the distance and number of synapses).

We thank the reviewer for pointing out these very useful studies. We have taken on board this suggestion for a future model of the mouse barrel cortex.

(2) The authors mention that in the model, the response of the main activated downstream area was confined to L6. Is this consistent with in vivo observations? Additionally, is there any in vivo characterization of the distance dependence of spiking correlation to validate Figure 8I?

We are not aware of data confirming the propagation of activity to downstream areas being confined to layer 6 but have considered the connectivity further between these two regions on page 27, as well as studying this further in follow up work:

“Stable propagation of evoked activity through mid-range connectivity only emerges in nbS1 at predicted in vivo firing rates

We repeated the previous single whisker deflection evoked activity experiment in the full model, providing a synchronous thalamic input into the forelimb sub-region (S1FL; Figure 8E; Video 8 & 9). Responses in S1FL were remarkably similar to the ones in the seven column subvolume, including the delays and decays of activity (Figure 8F). However, in addition to a localized primary response in S1FL within 350μm of the stimulus, we found several secondary responses at distal locations (Figure 8E; Video 9), which was suggestive of selective propagation of the stimulus-evoked signal to downstream areas efferently connected by mid-range connectivity. The response of the main activated downstream area (visible in Figure 8E) was confined to L6 (Figure 8G). In a follow up study using the model to explore the propagation of activity between cortical regions (Bolaños-Puchet and Reimann, 2024), it is described how the model contains both a feedforward projection pattern, which projects to principally to synapses in L1 & L23, and a feedback type pattern, which principally projects to synapses in L1 & L6. On visualizing the innervation profile from the stimulated hexagon to the downstream hexagon we can see that we have stimulated a feedback pathway (Figure S16)”

With referenced Figure S16 on page 85:

We did find in vivo evidence of similar layer-wise and distance dependence of correlations in the somatosensory cortex discussed on page 27 of the Results:

“The distance dependence of correlations followed a similar profile to that observed in a dataset characterizing spontaneous activity in the somatosensory cortex (Reyes-Puerta et al., 2015a) (compare red line in Figure 8I with Figure S16). In the in vivo dataset spiking correlation was also low but highest in lower layers, with short “up-states” in spiking activity constrained to L5 & 6 (see Figure 1E,F in (Reyes-Puerta et al., 2015a)). In the model, they are constrained to L6.”

With Figure S16a on page 85 showing the distance dependence of correlations in the anaesthetized barrel cortex during spontaneous activity (digitization from the reference paper):

(3) Across the figures, activity is averaged across neurons within layers and E or I cell types, with a limited description of single-cell type and single-cell responses. Were there any predictions regarding the responses of particular cell types that significantly differ from others in the same layer? Such predictions could be valuable for future investigations and could showcase the advantages of a data-driven, biophysically detailed model.

We thank the review for this comment. In addition to new analyses at higher granularity addressed in other comments, we have added the following comparison of stimulus-evoked membrane potential dynamics in different subpopulations for the original connectome and SM-connectome in Figure 7 on page 24.

This gave interesting results discussed in a new subsection on page 26:

“EM targeting trends hyperpolarize Sst+ and HT3aR+ late response, and disinhibit L5/6 E

Studying somatic membrane potentials for different subpopulations in response to whisker deflections shows that PV+, L23E and L4E subpopulations are largely unaffected in the SM-connectome (Figure 7E). Interestingly, Sst+ and 5HT3aR+ subpopulations show a strong hyperpolarization in the late response that isn’t present in the original connectome. Interestingly, this corresponds with a stronger late response in L5/6 E populations, which could be caused by disinhibition due to the Sst+ and 5HT3aR+ hyperpolarization. This could be explored further in follow up studies using our connectome manipulator tool (Pokorny et al., 2024).”

(4) 2.4: Are there caveats to assuming the OU process as a model for missing inputs? Inputs to the cortex are usually correlated and low-dimensional (i.e., communication subspace between cortical regions), but the OU process assumes independent conductance injection. Can (weakly) correlated inputs give rise to different activity regimes in the model? Can you add a discussion on this?

We agree with the reviewer that there are caveats to assuming an OU process for the model of missing inputs and have added the following to the Discussion on page 31:

“The calibration framework could optimize per population parameters for other compensation methods, whilst still offering an interpretable spectrum of firing rate regimes at different levels of PFR. For example, more realistic compensation schemes could be explored which introduce a) correlations between the inputs received by different neurons and b) compensation distributed across dendrites, as well as at the soma. We predict that such changes would make spontaneous activity more correlated at the lower spontaneous firing rates which supported in vivo like responses (PFR : 0.1 − 0.5), which would in turn make stimulus-responses more noise correlated.”

(5) 2.6: The network structure is well characterized in the companion paper, where the authors report that correlations in higher dimensions were driven by a small number of neurons with high participation ratios. It would be interesting to identify which cell types exhibit high node participation in high-dimensional simplices and examine the spiking activity of cells within these motifs. This could generate testable predictions and inform theoretical cell-type-specific point neuron models for excitatory/inhibitory balanced networks and cortical processing.

We thank the reviewer for this suggestion. We have added two supplementary figures to address this suggestion, which are discussed in the Results on Page 16:

“Additionally, we studied the structural effect on the firing rate (here measured as the inverse of the inter-spike interval, ISI, which can be thought of as a proxy of non-zero firing rate). We found that for the connected circuit, the firing rate increases with simplex dimension; in contrast with the disconnected circuit, where this relationship remains flat (see Figure S6 red vs. blue curves and Methods).

This also demonstrates high variability between neurons, in line with biology, both structurally (Towlson et al., 2013; Nigam et al., 2016) and functionally (Wohrer et al., 2013; Buzs´aki and Mizuseki, 2014). We next identified the cell types that are overexpressed in the group of neurons that have the 5% highest values of node participation across dimensions (Figure S7). This could inform theoretical point neuron models with cell-type specificity, for example. We found that while in dimension one (i.e., node degree) this consists mostly of inhibitory cells, in higher dimensions the cell types concentrate in layers 4, 5 and 6, especially for TPC neurons. This is in line with our structural layer-wise findings in Figure 8B in Reimann et al. (2024).”

Which reference new Figures S6 and S7:

With the methodology for S6 described on page 49 of the Methods:

“For any numeric property of neurons, e.g., firing rate, we evaluate the effect of dimension on it by taking weighted averages across dimensions. That is for each dimension k, we take the weighted average of the property across neurons where the weights are given by node participation on dimension k. More precisely, let N be the number of neurons and −→V ∈ RN, be a vector of a property on all the neurons e.g., the vector of firing rates. Then in each dimension k we compute

Where is the vector of node participation on dimension k for all neurons and ・ is the dot product.

To measure the over and underexpression of the different m-types among those with the highest 5% of values of node participation, we used the hypergeometric distribution to determine the expected distribution of m-types in a random sample of the same size. More precisely, for each dimension k and m-type m, let Ntotal be the total number of neurons in the circuit, Nm be the number of neurons of m-type m in the circuit, Ctop be the number of neurons with the highest 5% values of node participation in dimension k, Cm the number of neurons of mtype m among these, and let P = hypergeom(Ntotal</sub<,Nm,Ctop) be the hypergeometric distribution.

By definition, P(x) describes the probability of sampling x neurons of m-type m in a random sample of size Ctop. Therefore, using the cumulative distribution F(x) = P(Counts ≤ x), we can compute the p-values as follows:

Small values indicate under and over representation respectively….”

Minor:

(1) Since the previous model was published in 2015, the neuroscience field has seen significant advancements in single-cell and single-nucleus sequencing, leading to the clustering of transcriptomic cell types in the entire mouse brain. For instance, the Allen Institute has identified ~10 distinct glutamatergic cell types in layer 5, which exceeds the number incorporated into the current model. Could you discuss 1) the relationship between the modeled me-types and these transcriptomic cell types, and 2) how future models will evolve to integrate this new information? If there are gaps in knowledge in order to incorporate some transcriptome cell types into your model, it would be helpful to highlight them so that efforts can be directed toward addressing these areas.

We thank the reviewer for this suggestion, particularly the idea to describe what types of data would be valuable towards improving the model in future. We have added the following to the Discussion on page 33:

“In our previous work (Roussel et al., 2023) we linked mouse inhibitory me-models to transcriptomic types (t-types) in a whole mouse cortex transcriptomic dataset (Gouwens et al., 2019). This can provide a direct correspondence in future large-scale mouse models. As we model only a single electrical type for pyramidal cells there is no one-to-one correspondence between our me-models and the 10 different pyramidal cell types identified there. We are not currently aware of any method which can recreate the electrical features of different types of pyramidal cells using only generic ion channel models. To achieve the firing pattern behavior of more specific electrical types, usually ion channel kinetics are tweaked, and this would violate the compartmentalization of parameters. In future we hope to build morpho-electric-transcriptomic type (met-type) models by selecting gene-specific ion channel models (Ranjan et al., 2019, 2024) based on the met-type’s gene expression. Data specific to different neuron sections (i.e. soma, AIS, apical/basel dendrites) of different met-types, such as gene expression, distribution of ion channels, and voltage recordings under standard single cell protocols would be particularly useful.”

(2) For the optogenetic manipulation, it would be interesting if the model could reproduce the paradoxical effects (for example, Mahrach et al. reported paradoxical effects caused by PV manipulation in S1; https://pubmed.ncbi.nlm.nih.gov/31951197/). This seems a more relevant and non-trivial network phenomenon than the V1 manipulation the authors attempted to replicate.

We thank the reviewer for this valuable idea. Indeed, our model is able to reproduce paradoxical effects under certain conditions. We added the following new supplementary Figure S12 demonstrating this finding (black arrows).

Which we discuss in the Results on page 22:

“However, at high contrasts, we observed a paradoxical effect of the optogenetic stimulation on L6 PV+ neurons, reducing their activity with increasing stimulation strength (Figure S12B; cf. Mahrach et al. (2020)). This effect did not occur under grey screen conditions (i.e., at contrast 0.0) with a constant background firing rate of 0.2 Hz or 5 Hz respectively (not shown). The individual…”

and added to the Discussion on page 32:

“Also, we predicted a paradoxical effect of optogenetic stimulation on L6 PV+ interneurons, namely a decrease in firing with increased stimulus strength. This is reminiscent of the paradoxical responses found by Mahrach et al. (2020) in the mouse anterior lateral motor cortex (in L5, but not in L2/3) and barrel cortex (no layer distinction) respectively. While Mahrach et al. (2020) conducted their recordings in awake mice not engaged in any behavior, we observed this effect only when drifting grating patterns with high contrast were presented. Nevertheless, consistent with their findings, we found the effect only in deep but not in superficial layers, and only for PV+ interneurons but not for PCs. Our model could therefore be used to improve the understanding of this paradoxical effect in follow up studies. These examples demonstrate that the approach of modeling entire brain regions can be used to further probe the topics of the original articles and cortical processing.”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

My specific comments are in the Public Review. The summarizing point is that this is a sprawling paper, and it is easy for readers to get confused. Focusing on specific connections between known functional properties and findings in this model, especially for the full-scale model, will be helpful.

We thank the reviewer for this comment and for their related recommendation (4) below, and have added subheadings through-out the results.

Reviewer #2 (Recommendations for the authors):

(1) P4. What are the 10 free parameters?

We thank the reviewer for pointing out that it would be useful to summarize the 10 parameters at this stage of the text, and have adjusted the sentence to:

“As a result, the emerging in-vivo like activity is the consequence of only 10 free parameters representing the strength of extrinsic input from other brain regions into 9 layer-specific excitatory and inhibitory populations, and a parameter controlling the noise structure of this extrinsic input.”

(2) Table 1 and S1 are extremely useful. Could you provide a table summarizing the major assumptions or gaps in the model, their potential influence on the results, and possible ways to collect data that could support or challenge these assumptions? Currently, this information is scattered throughout the manuscript.

We thank the reviewer for this very useful suggestion and have added a Table S8 on page 68:

(3) Figure 4F is important, but the legend is unclear. What is the unit on the x-axis? The values seem too large to represent per-neuron measurements.

Thank you to the reviewer for raising this. Indeed the values are estimated mean numbers of missing number synapses per neuron by population. Such numbers are difficult to estimate but we have further discussed our rationale, justification and consideration of whether these numbers are accurate in the Results, as follows:

“Heterogeneity in synaptic density within and across neuron classes and sections makes estimating the number of missing synapses challenging (DeFelipe and Fariñas, 1992). Changing the assumed synaptic density value of 1.1 synapses/μm would only change the slope of the relationship, however. Estimates of mean number of existing and missing synapses per population were within reasonable ranges; even the larger estimate for L5 E (due to higher dendritic length; Figure S3) was within biological estimates of 13,000 ± 3,500 total afferent synapses (DeFelipe and Fariñas, 1992).”

This text references the new supplementary Figure S3:

Moreover, these numbers represent the number of synapses, rather than the number of connections. The number of connections is usually used for quantifications such as indegree, and are usually much lower.

We have also updated the caption and axis labels of the original figure:

(4) Including additional subsections or improving the indexing in the Results section could be beneficial. In its current format, it's difficult to distinguish where the model description ends and where the validation begins. Some readers may want to focus more on the validation than other parts, so clearer segmentation would improve readability.

We have addressed this comment with the opening comment in the authors “Recommendations for authors”.

(5) P4. 2nd paragraph. Original vs rewired connectome. The term "rewired connectome" may give the impression that it refers to an artificial manipulation rather than a modification based on the latest data. It might be helpful to use a different term (e.g., SM-connectome as described later in the paper?).

We have adjusted the text in the introduction:

“Additionally, we generated a new connectome which captured recently characterized spatially-specific targeting rules for different inhibitory neuron types (Schneider-Mizell et al., 2023) in the MICrONS electron microscopy dataset (MICrONS-Consortium et al., 2021), such as increased perisomatic targeting by PV+ neurons, and increased targeting of inhibitory populations by VIP+ neurons. Comparing activity to the original connectome gave predictions about the role of these additional targeting rules.”

(6) Figures 7 B, C, D: what is v1/v2? Original vs SM-Connectome?

We thank the reviewer for noticing this and have corrected the figure to use “Orig” and “SM” consistent with the rest of the figure.

(7) Page 23, 2.10: what is phi?

We thank the reviewer for noticing this inconsistency with the earlier text, and have updated the text to read: “Particularly, we ran simulations for PF R ∈ [0.1, 0.15, ..., 0.3] using the OU para-maters calibrated for the seven column subvolume for [Ca2+] = 1.05 mM and ROU = 0.4.”

Version 2

Abstract

Summary

Cortical dynamics underlie many cognitive processes and emerge from complex multi-scale interactions, which are challenging to study in vivo. Large-scale, biophysically detailed models offer a tool which can complement laboratory approaches. We present a model comprising eight somatosensory cortex subregions, 4.2 million morphological and electrically-detailed neurons, and 13.2 billion local and mid-range synapses. In silico tools enabled reproduction and extension of complex laboratory experiments under a single parameterization, providing strong validation. The model reproduced millisecond-precise stimulus-responses, stimulus-encoding under targeted optogenetic activation, and selective propagation of stimulus-evoked activity to downstream areas. The model’s direct correspondence with biology generated predictions about how multiscale organization shapes activity; for example, how cortical activity is shaped by high-dimensional connectivity motifs in local and mid-range connectivity, and spatial targeting rules by inhibitory subpopulations. The latter was facilitated using a rewired connectome which included specific targeting rules observed for different inhibitory neuron types in electron microscopy. The model also predicted the role of inhibitory interneuron types and different layers in stimulus encoding. Simulation tools and a large subvolume of the model are made available to enable further community-driven improvement, validation and investigation.

Introduction

Neuroscience aims to characterize the dynamics of the brain and understand how they emerge as a product of anatomy and physiology. This is challenging however, due to the complexity of the brain’s multi-scale organization and its study in situ. Large-scale, data-driven, biophysically-detailed models (Markram et al., 2015; Billeh et al., 2020) offer a tool which can complement laboratory investigation. This “bottom-up” approach combines detailed constituent models into a 3D model of a given brain volume, thus consolidating disparate, multi-scale data sources. Such models aim to provide a general tool for multi-scale investigation; access and manipulation of any model component enables predictions about how these components combine to shape emergent dynamics. With increasing biological data, model refinement and validation must be continuous. This necessitates 1) open sourcing of models with high quality software tools for iterative, community-driven refinement and validation, and 2) rigorous methodology for model parameterization and validation, such that model iterations can be meaningfully compared. Here and in our companion paper (Reimann et al., 2024) we present our methodology for parameterizing the anatomy and physiology of a cortical model, and validating its emergent dynamics. This enabled predictions about how cortical activity is shaped by high-dimensional connectivity motifs in local and mid-range connectivity, and spatial targeting rules by inhibitory subpopulations. Elsewhere, this has allowed the presented model to be used to study the formation of cell assemblies (Ecker et al., 2024b), functional synaptic plasticity (Ecker et al., 2024a), propagation of activity between cortical areas (Bolaños-Puchet and Reimann, 2024) the role of non-random connectivity motifs on network activity (Pokorny et al., 2024) and reliability (Egas Santander et al., 2024), the composition of high-level electrical signals such as the EEG (Tharayil et al., 2024), and how spike sorting biases population codes (Laquitaine et al., 2024).

Specifically, we built and validated a model of the entire non-barrel primary somatosensory cortex (nbS1) comprising eight subregions. Whilst our previous data-driven, biophysically detailed model (Markram et al., 2015) provided insights at the scale of a single cortical column (Reimann et al., 2013, 2017, 2022; Nolte et al., 2019, 2020; Newton et al., 2021), the new model is ∼140 times larger and to our knowledge offers the first simulations of in vivo-like spontaneous and stimulus-evoked activity in a biophysically detailed cortical model with interregion connectivity. In the companion paper, we introduce the anatomical model (Fig. 1, Step 1), describing how neuron morphologies were placed within an atlas-based geometry and connected through local and mid-range synapses. Here, we describe our improved techniques to model and validate the electrical properties of neurons and synapses (Figure 1, Steps 2 & 3 (Reva et al., 2023; Barros-Zulaica et al., 2019)), and to compensate for input from missing brain areas (Figure 1, Steps 4). These improvements enabled enhanced validation of emerging in vivo-like activity (Figure 1, Steps 5), including the reproduction and extension of five published studies in rodent sensory cortex under a single in vivo-like regime (Figure 1, Steps 6).

Overview of the physiology and simulation workflow.

1. Anatomical model: Summary of the anatomical nbS1 model described in the companion paper. 2. Neuron physiology: Neurons were modeled as multi-compartment models with ion channel densities optimised using previously established methods and data from somatic and dendritic recordings of membrane potentials in vitro. 3. Synaptic physiology: Models of synapses were built using previously established methods and data from paired recordings in vitro. 4. Compensation for missing synapses: Excitatory synapses originating from outside nbS1 were compensated with noisy somatic conductance injection, parameterized by a novel algorithm. 5. In vivo-like activity: We calibrated an in silico activity regime compatible with in vivo spontaneous and stimulus-evoked activity. 6. In silico experimentation: Five laboratory experiments were recreated. Two were used for calibration and three of them were extended beyond their original scope. 7. Open Source: Simulation software and a seven column subvolume of the model are available on Zenodo (see data availability statement). Data generalisations: Three data generalisation strategies were employed to obtain the required data. Left: Mouse to rat, middle: Adult to juvenile (P14) rat, right: Hindlimb (S1HL) and barrel field (S1BF) subregions to the whole nbS1. Throughout the figure, the corresponding purple icons show where these strategies were used.

Validating that network activity emerges from the same interactions driving in vivo dynamics, requires 1) an approach to tackle the large parameter space without overfitting and 2) comparison of emerging dynamics with laboratory experiments. With respect to, we strongly prefer parameterization with directly measured quantities over fitting parameters to yield the correct emerging activity. Additionally, where fitting is applied, we adhere to a principle of compartmentalization of parameters. That is, once a parameter has been parameterized at one biological level, it is no longer a free parameter at a higher level. For example, the maximal conductance of a synapse is fit to biological amplitudes of postsynaptic potentials, but is then never updated in the process of reaching in-vivo like population activity. If a connection is valid on the single-cell level, its contribution at the population level should be equally valid. As a result, the emerging in-vivo like activity is the consequence of only 10 free parameters representing the strength of extrinsic input from other brain regions into 9 layer-specific excitatory and inhibitory populations, and a parameter controlling the noise structure of this extrinsic input. These parameters were fitted using a novel methodology presented here and were in accordance with the mean number of missing synapses for each population (Felleman and Van Essen, 1991; Harris et al., 2019; Gao et al., 2022).

With respect to 2), we conducted the following validations: Spontaneous activity reproduced layer-wise in vivo firing rates (or a specified proportion of in vivo firing rates to account for in vivo recording bias (Wohrer et al., 2013)), varied along a spectrum of asynchronous to synchronous activity, exhibited spatially structured fluctuations and produced long-tailed firing rate distributions with sub 1 Hz peaks as in vivo (Wohrer et al., 2013; Buzsáki and Mizuseki, 2014). Under the same parameterization, the model reproduced precise millisecond dynamics of layer-wise populations in response to simple stimuli. These initial validations demonstrated that the model was in a more accurate regime compared to Markram et al. (2015) – an essential step before testing more complex or larger-scale validations. For example, under the same parameterization we then observed selective propagation of stimulus-evoked activity to downstream areas, and reproduced and extended more complex experiments through accurate modeling of targeted optogenetic stimulation and lesions. Importantly, we highlight where emergent activity shows discrepancies with in vivo activity, to guide future data-driven model refinement.

The model generated a number of predictions (Table S1), including about the role of different layers in driving layer 2/3 stimulus responses and how inhibitory interneuron types encode contrast, synchronous and rate-coded information. Additionally, with access to the full structural connectome and tools for precisely editing it (Pokorny et al., 2024), we were able to make predictions about the relationship between structure and function. For example, we predict that an increase in the prevalence of non-random connectivity motifs towards deeper layers leads to a matched increase in spiking correlations, and that subregions more strongly innervated by mid-range connectivity have higher correlated activity locally. Additionally, we generated a new connectome which captured recently characterized spatially-specific targeting rules for different inhibitory neuron types (Schneider-Mizell et al., 2023) in the MICrONS electron microscopy dataset (MICrONS-Consortium et al., 2021), such as increased perisomatic targeting by PV+ neurons, and increased targeting of inhibitory populations by VIP+ neurons. Comparing activity to the original connectome gave predictions about the role of these additional targeting rules. For example, inhibitory populations were more strongly inhibited (increasingly towards central layers) and required more non-local drive to reach the firing rates observed in vivo. Evoked responses increased and decreased in superficial and deeper excitatory populations respectively, suggesting layer-specific roles of the more specific inhibitory targeting.

To provide a framework for further studies and integration of experimental data, the full model is made available with simulation tools, as well as a smaller subvolume with the optional new connectome capturing inhibitory targeting rules from electron microscopy. (Figure 1, Step 7). The detailed modeling approach provides a one-to-one correspondence with most types of experimental data, allowing different datasets to be readily integrated. Due to the incredible speed of discovery in neuroscience an integrative model will always be lagging behind the latest available data. We believe the solution is to provide a scientifically solid, validated model with clearly characterized strengths and weaknesses, along with the tools to advance or customize it for individual projects. We have therefore also made our tools for building and improving the model openly available: www.github.com/BlueBrain.

Results

The companion paper describes the full anatomical nbS1 model containing 4.2M morphological neuron models, connected through 9.1B local synapses and 4.1B mid-range synapses (Reimann et al., 2024). Each neuron is modeled as a multi-compartmental model belonging to one of 60 morphological types (m-types; Figure 2A1), and is either an instance or statistical variant of an exemplar in a pool of 1,017 morphological reconstructions. Neurons were placed (orientated towards the surface) within an atlas-based geometry (Figure 2A2) based on estimated layer-wise density profiles of different morphological types. Local connectivity is based on axo-dendritic overlap with neighbouring neurons, whilst mid-range connectivity combines data on interregion connectivity and laminar innervation profiles. Thalamocortical afferents were also modeled based on laminar innervation to the barrel cortex. To simulate emergent activity it is necessary to model and validate the electrical properties of these neurons and synapses (Figure 1 Steps 2 & 3). These modeling steps are based on published methods and data sources (Table 1), and are summarized first. The method for compensating for missing synapses and the remaining results are then described (Figure 1 Steps 4-6).

Modeling and validation of neuron physiology.

A1: Example excitatory neuron morphologies. See companion paper for exemplar morphologies of all morphological types. A2: Example morphologies placed within the atlas-based volume. Axons shown in grey. B: Optimized conductance densities for two exemplary e-types. B1: cADpyr e-type on L5 TPC:A m-type, B2: bSTUT and cSTUT e-types on L5 NBC m-type. (The L5 NBC m-type is combined with more e-types than the two shown, see panel D.) Morphologies were visualized with NeuroMorphoVis (Abdellah et al., 2018). Neurite diameters are enlarged (x3) for visibility. Soma and dendrites in black, axon in red. C: e-types used in the model. As used in Markram et al. (2015) and similar to the Petilla terminology (Ascoli et al., 2008)).

D: me-type composition. Heatmap shows the proportion of e-types for each m-type. Each me-type is assigned to one of the three I subpopulations. Assignments depending not only on m-type are highlighted by a green box. Particularly, for BTC and DBC m-types, the cACint e-type belongs to the Sst+ subpopulation. E: Validation of dendritic physiology of all L5 TTPCs. Panel reproduced from Figure 5B and C of Reva et al. (2023). E1: Validation of back-propagating action potential (bAP) amplitude (i.e., the dependence of bAP amplitude on distance from the soma) for basal (green) and apical (blue) dendrites. Reference data (in red) comes from Stuart and Sakmann (1994); Larkum et al. (2001) (apical) and Nevian et al. (2007) (basal). Lines show exponential fits for the in silico (green and blue) and in vitro (red) data. Color bar indicates dendritic diameter. E2: Validation of EPSP attenuation. Reference data (in red) comes from Berger et al. (2001) (apical) and Nevian et al. (2007) (basal). Lines and colorbar same as in E1.

Improved modeling and validation of neuron physiology

Similarly to Markram et al. (2015), electrical properties of single neurons were modeled by optimizing ion channel densities in specific compartment-types (soma, axon initial segment (AIS), basal dendrite, and apical dendrite) (Figure 2B) using an evolutionary algorithm (IBEA; Van Geit et al., 2016) so that each neuron recreates electrical features of its corresponding electrical type (e-type) under multiple standardized protocols. Compared to Markram et al. (2015), electrical models were optimized and validated using 1) additional in vitro data, features and protocols, 2) ion channel and electrophysiological data corrected for the liquid junction potential, and 3) stochastic channels (StochKv3) now including inactivation profiles. The methodology and resulting electrical models are described in Reva et al. (2023) (see Methods), and generated quantitatively more accurate electrical activity, including improved attenuation of excitatory postsynaptic potentials (EPSPs) and back-propagating action potentials. Electrical features included firing properties (e.g. spike frequency, inter-spike interval), action potential waveforms (e.g. fall and rise time, width) and passive properties (e.g. input resistance). The optimization was performed for a subset of neuron models. The resulting ion channel densities were generalized to other neuron models of the same e-type. For excitatory neurons such generalization was only made to neurons within the same layer. The resulting electrical activity of each neuron was validated against the corresponding electrical features, including the characteristic firing properties of the 11 e-types (Figure 2C). Model neurons with a mean zscore (over the electrical features) further than two standard deviations from the experimental mean were discarded. The new neuron models saw a 5-fold improvement in generalizability compared to Markram et al. (2015) (Reva et al., 2023).

For each of the 60 morphological types (m-types), the corresponding fractions of e-types were determined from experimental data as in Markram et al. (2015), resulting in 208 morpho-electrical types (me-types; Figure 2D). Correspondence with predominant expression of biological markers PV (parvalbumin), Sst (somatostatin) or 5HT3aR (serotonin receptor 3A) was determined based on me-type, as previously done in Markram et al. (2015) (Figure 2D). Finally, for layer five thick-tufted pyramidal cell (L5 TTPC) morphologies, we found that dendritic electrical features, namely the attenuation of back propagating action potentials and EPSPs, reproduced experimental measurements (Figure 2E, Reva et al., 2023).

The required data to constrain region-specific me-type distributions was not available when building the model; to our knowledge Yao et al. (2023) is the first study that explored the question systematically (in mouse cortex). The atlas-based geometry does however impose heterogeneity in morphologies that e-types are paired with: only morphologies that are in line with the region-specific local curvature of the region are used throughout the volume. We demonstrate the effect of geometry on morphological composition in the companion paper.

Improved modeling and validation of synaptic physiology

The biological realism of synaptic physiology was improved relative to Markram et al. (2015) using additional data sources and by extending the stochastic version of the Tsodyks-Markram model (Tsodyks and Markram, 1997; Markram et al., 1998; Fuhrmann et al., 2002; Loebel et al., 2009) to feature multi-vesicular release, which in turn improved the accuracy of the coefficient of variations (CV; std/mean) of postsynaptic potentials (PSPs) as described in Barros-Zulaica et al. (2019) and Ecker et al. (2020). The model assumes a pool of available vesicles that is utilized by a presynaptic action potential, with a release probability dependent on the extracellular calcium concentration ([Ca2+]o; Ohana and Sakmann, 1998; Rozov et al., 2001; Borst, 2010). Additionally, single vesicles spontaneously release as an additional source of variability with a low frequency (with improved calibration relative to Markram et al. (2015)). The utilization of vesicles leads to a postsynaptic conductance with bi-exponential kinetics. Short-term plasticity (STP) dynamics in response to sustained presynaptic activation are either facilitating (E1/I1), depressing (E2/I2), or pseudo-linear (I3). E synaptic currents consist of both AMPA and NMDA components, whilst I currents consist of a single GABAA component, except for neurogliaform cells, whose synapses also feature a slow GABAB component. The NMDA component of E synaptic currents depends on the state of the Mg2+ block (Jahr and Stevens, 1990), with the improved fitting of parameters to cortical recordings from Vargas-Caballero and Robinson (2003) by Chindemi et al. (2022).

The workflow for determining a dense parameter set for all synaptic pathways, starting with sparse data from the literature, is described for the use case of hippocampal CA1 in Ecker et al. (2020) and briefly in the Methods. We combined data sources used in Markram et al. (2015), with a large number of recent data sources (Qi and Feldmeyer, 2016; Barros-Zulaica et al., 2019; Yang et al., 2020, 2022). The resulting pathway-specific parameters are listed in Tables S2, S3, S4, the most common short-term dynamics are depicted in Figure 3A1-2, and the assignment of STP profiles to different pathways are shown in Figure 3A3. Postsynaptic potential amplitudes and their CV closely matched their biological counterparts (r = 0.99, n = 27; Figure 3B1; Table S5 and r = 0.63, n = 10; Figure 3C1; Table S6, respectively). The dense parameter set also allowed prediction of PSP amplitudes and CVs for all cortical pathways (Figure 3B2, C2). The frequency of miniature postsynaptic currents (mPSCs) were also in line with in vitro measurements (r = 0.92, n = 5; Figure 3D; Table S7).

Modeling and validation of synaptic physiology.

Caption on the following page. A1: Exemplary pair of L5 TTPCs (visualized with NeuroMorphoVis (Abdellah et al., 2018)). Presynaptic cell in gray, postsynaptic cell in red, synapses between them in purple. Neurite diameters are enlarged (x3) for visibility and axons were cut to fit into the figure. Pre- and postsynaptic voltage traces on the top right. A2: Exemplary postsynaptic traces with different STP profiles. A3: Assignment of STP profiles to viable pathways. (Pathways were considered viable if there were at least 10 connections in all eight subregions of the model.) B1: Validation of first PSP amplitudes (see also Table S5). Dashed gray line represents perfect correlation between experimental and model values. Error bars show one standard deviation (also for C1, D and F). B2: Predicted PSP amplitudes of all viable pathways in the circuit. Postsynaptic cells were held at -70 mV using an in silico voltage-clamp. Means were calculated over 100 pairs of neurons with 35 repetitions each. C1 and C2: same as B1 and B2, but showing the CV of the first PSP amplitude (corresponding Table is S6). D: Validation of mPSC frequencies (see also Table S7). E: Location of synapses from VPM fibers (purple) and POm fibers (red) on 38 neurons (dark gray) in a 5 µm radius column (visualized with BioExplorer). F: Validation of thalamocortical EPSP amplitudes as in B1. The four pathways used for the validation are marked with a black rectangle on H1 to its right. G: EPSP latencies (time from presynaptic spike to the rise to 5% of peak EPSP amplitude in the postsynaptic trace). H1 Left: mean VPM evoked EPSP amplitudes on postsynaptic cell types (over 50 pairs). Right: Comparison of normalized in silico amplitudes (normalized by L4 E as in Sermet et al., 2019) to in vitro reference data from Sermet et al. (2019). Heatmap shows model minus reference values, thus positive values indicate a higher normalized EPSP amplitude in our model than in the reference experimental dataset. H2: same as H1 but for POm (normalized by L5 E as in Sermet et al., 2019).

Improved modeling and validation of thalamocortical projections

The anatomical model includes fibers from the thalamus, based on fibers projecting to the barrel cortex from the ventral posteriomedial (VPM) and posteriormedial (POm) thalamic nuclei. These fibers make synaptic contacts within a radius of the fiber proba-bilistically based on laminar innervation profiles (see Reimann et al., 2024; Figure 3E). Compared to the previous model (Markram et al., 2015), POm projections are added and the physiology of synapses from VPM improved (see Methods; Figure 3E, F). Latencies of layer-wise EPSPs increase with distance from the thalamus (Figure 3G). Additionally, thalamocortical EPSP amplitudes normalized relative to a single population were compared to normalized EPSPs in response to optogenetic stimulation targeting bundles of thalamic fibers (Sermet et al., 2019). This provided contrasting insights, however. For example, whilst VPM to L6 I EPSPs match the initial validation data (Figure 3F), VPM to L6 PV+ responses appear too strong relative to other populations (Figure 3H1). The results suggest that the model’s POm to L5 E pathway is too weak, when compared to other POm to E and all POm to PV+ pathways (Figure 3H2, right).

Defining subvolumes and populations

To enable smaller simulations and targeted analyses, we defined standardized partitions comprising all neurons (and their connections) contained in a subvolume of the full nbS1 model. We decomposed the model into full-depth hexagonal prisms of a certain diameter that we call hexagonal subvolumes. These are slightly curved, following the geometry of the cortex, and have an intact layer structure. When the diameter of the hexagons is 520 µm, comparable to the size of the single cortical column model from (Markram et al., 2015), we call these subvolumes columns. Taken together, a central column and the six columns surrounding it define a seven column subvolume (Figure 4A). Additionally, when we separately analyze the E and I neurons in different layers, we refer to them as neuron populations of the model or a subvolume. Groups of I neurons that predominantly express either PV, Sst or 5HT3aR markers, are referred to as inhibitory subpopulations.

Spontaneous activity: calibration, in vivo-like dynamics, linking structure to function.

A: Seven column subvolume. B: OUµ and OUσ of somatic conductance injection (B1) parameterized by population (B2) determine FRs (B3). Target FRs equal to in vivo firing rates multiplied by PF R. Inset: Table summarising meta-parameters. C: Target PF R (x-axis) plotted against the observed PF R (y-axis) for each E (red) and I (blue) population after calibration. The number of sides of the shape indicates the layer (triangle used for L2/3). D: Euclidean distance between target and observed PF R values (over populations) decreases over iterations for two meta-parameter combinations (upper). Values after final iteration shown for all meta-parameter combinations (lower; dashed line shows termination condition). E: (left) Firing rates, (centre) mean conductance injection and (right) OUµ by population for the non-bursting simulations (1 line per simulation; E and I values separated). F: Estimated mean number of missing number synapses per neuron by population vs. the mean conductance injection, averaged over all combinations, for each population. Markers as in C. G1: Max normalised histograms for two meta-parameter combinations (5 ms bin size, 1σ Gaussian smoothing). G2: Effect of meta-parameters on correlation between E and I histograms (5 ms bin size, 1σ Gaussian smoothing; central column). G3: Fourier analysis of spontaneous activity for the 59 non-bursting simulations. H: Activity in flat space over seven consecutive 100 ms windows. Activity smoothed (Gaussian kernel, σ = 1pixel). I: Left: Correlation r-value between the histograms of layer-wise E and I populations for the 59 non-bursting simulations. Right: RC/U by population for the 59 non-bursting simulations. J: Top: The mean node participation of E and I populations in each layer for dimension one (left) and six (right) vs. RC/U for the parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4 and PF R = 0.3. Markers as in C. Line shows linear fit. Bottom: Correlation r-values between mean node participation and RC/U when neurons with highest node participation are not included in the calculation of mean node participation (via a sliding threshold).

Compensating for non-modeled brain regions

Algorithm efficiently finds spectrum of network regimes compatible with in vivo whilst accounting for unknown in vivo recording bias

Initial simulations produced no activity, and it was necessary to compensate for missing excitatory input from neurons external to the model. We estimated that the number of missing synapses for the full nbS1 and seven hexagon subvolume as ∼ two and seven times the number of internal synapses, based on Oh et al. (2014). We model the effect of these missing synaptic inputs explicitly as time-varying and statistically independent somatic conductance injections, using Ornstein-Uhlenbeck (OU) processes that mimic aggregated random background synaptic inputs (Destexhe et al., 2001; Figure 4B1; see also Discussion). The mean (OUµ) and standard deviation (OUσ) are defined as a percentage of an individual cell’s input conductance at its resting membrane potential. The choice of OUµ and OUσ for different populations and [Ca2+]o (Figure 4B2), determine the model’s emergent dynamics and mean firing rates (Figure 4B3). Initial simulations showed that using the same value of OUµ and OUσ for all neurons makes some populations highly active whilst others are silent, or produces network-wide bursts. Finding population-specific OU parameters which produce stable in vivo-like activity is challenging, however, due to the non-linearity and computational cost of simulations.

We developed an algorithm that efficiently calibrates population-specific OU parameters, for different values of [Ca2+]o, which affects the strength and reliability of synaptic transmission (see “Synaptic Physiology”). Additionally, it fixes the ratio OUσ/OUµ, denoted ROU, across all neurons to specific values representing the amount of noise in the extrinsic inputs (Figure 4B2). As extracellularly-derived firing rates are known to be severely overestimated to an unknown degree (Olshausen and Field, 2006; Wohrer et al., 2013; Buzsáki and Mizuseki, 2014), the algorithm finds parameters which produce realistic inter-population firing rate ratios at different global levels of activity (Figure 4B3). Specifically, to account for uncertainty in the firing rate bias during spontaneous activity from extracellular spike sorted recordings it targets mean spontaneous firing rates of neuron populations that are a constant fraction PF R of in vivo reference values, for 10 PF R values between 0.1 and 1. We refer to [Ca2+]o, ROU and PF R as meta-parameters. The initial calibration was made for the seven hexagonal subvolume, and later generalised to the full nbS1 model.

The algorithm was first run for [Ca2+]o = 1.1 mM, ROU = 0.4. Target firing rates for the 10 PF R values were reached (Figure 4C) after five iterations of 10 simulations of 6.5 seconds (Figure 4D, left upper). Only three iterations were needed when calibration began using previously calibrated parameters for different ROU and [Ca2+]o (Figure 4D, right upper). This allowed us to parameterize extrinsic input for a range of combinations of the three meta-parameters (Figure 4D, lower & 4E, left) within biologically relevant ranges ([Ca2+]o: 1.05 mM - 1.1 mM (Jones and Keep, 1988; Massimini and Amzica, 2001; Amzica et al., 2002; Gonzalez et al., 2022); PF R: 0.1 - 1.0; ROU : 0.2 - 0.4 (Destexhe et al., 2001)). We found that the resulting mean conductance injections were highly population-specific (Figure 4E, centre). The range of OUµ values between populations was on average 5.2 times higher than the range of OUµ values for a single population across meta-parameter combinations (Figure 4E, right). This was expected, as the missing mid-range innervation is known to have specific and heterogeneous laminar profiles (Felleman and Van Essen, 1991; Harris et al., 2019; Gao et al., 2022).

Population-wise input compensation correlates with estimated number of missing synapses

We estimated the number of missing synapses per neuron assuming a total density of E synapses of 1.1 synapses/µm, based on mean spine densities (Larkman, 1991; Datwani et al., 2002; Kawaguchi et al., 2006), and subtracting the number of synapses present in the model (Figure S3). We confirmed a strong correlation between this measure and required conductance injection (Figure 4F). Heterogeneity in synaptic density within and across neuron classes and sections makes estimating the number of missing synapses challenging (DeFelipe and Fariñas, 1992). Changing the assumed synaptic density value of synapses/µm would only change the slope of the relationship, however. Estimates of mean number of existing and missing synapses per population were within reasonable ranges; even the larger estimate for L5 E (due to higher dendritic length; Figure S3) was within biological estimates of 13,000 ± 3,500 total afferent synapses (DeFelipe and Fariñas, 1992).

In vivo-like spontaneous activity

PF R, ROU and [Ca2+]o determine the properties of correlative spontaneous activity

Rasters and histograms of spontaneous activity for two meta-parameter combinations are shown in Figure 4G1 (and for all combinations in Video 1). The first simulation uses a higher ROU, lower PF R and lower synaptic transmission reliability ([Ca2+]o). In the second simulation, we found larger fluctuations, correlated between E and I populations across layers. Overall, increasing [Ca2+]o and PF R, and decreasing ROU, increases the amplitude of correlated fluctuations (Video 1; Figure 4G2) and increases power at lower frequencies (Figure 4G3). This supports the notion that decreasing ROU or increasing [Ca2+]o shifts the model along a spectrum from externally (noise) driven to internally driven correlated activity. Note that the most correlated meta-parameter combination [Ca2+]o: 1.1mM, ROU : 0.2, PF R: 1.0 produced network-wide “bursting” activity, which we define as highly synchronous all or nothing events (Video 1). Such activity, which may be characteristic of epileptic activity, can be studied with the model but is not the focus of this study.

We reemphasize that the [Ca2+]o, ROU and PF R meta-parameters account for uncertainty of in vivo extracellular calcium concentration, the nature of inputs from other brain regions and the bias of extracellularly recorded firing rates. Whilst estimates for [Ca2+]o are between 1.0 - 1.1mM (Jones and Keep, 1988; Massimini and Amzica, 2001; Amzica et al., 2002; Gonzalez et al., 2022) and estimates for PF R are in the range of 0.1 - 0.3 (Olshausen and Field, 2006), combinations of these parameters supporting in vivo-like stimulus responses in later sections will offer a prediction for the true values of these parameters. Both these later results and our recent analysis of spike sorting bias using this model (Laquitaine et al., 2024) predict a spike sorting bias corresponding to PF R ∼ 0.3, confirming the prediction of Olshausen and Field (2006).

Long-tailed population firing rate distributions with means ∼ 1 Hz

To study the firing rate distributions of different subpopulations and m-types, we ran 50s simulations for the meta-parameter combinations: [Ca2+]o: 1.05 mM, ROU : 0.4, PF R: 0.3, 0.7 (Figure S4). Different subpopulations showed different sparsity levels (proportion of neurons spiking at least once) ranging from 6.6 to 42.5%. Wohrer et al. (2013) considered in detail the biases and challenges in obtaining ground truth firing rate distributions in vivo, and discuss the wide heterogeneity of reports in different modalities using different recording techniques. They conclude that most evidence points towards longtailed distributions with peaks just below 1Hz. We confirmed that spontaneous firing rate distributions were long-tailed (approximately lognormally distributed) with means on the order of 1 Hz for most subpopulations. Importantly the layer-wise means were just below 1 Hz in all layers for the PF R = 0.3 meta-parameter combination. Moreover, our recent work applying spike sorting to extracellular activity using this meta-parameter combination found spike sorted firing rate distributions to be lognormally distributed and very similar to in vivo distributions obtained using the same probe geometry and spike sorter (Laquitaine et al., 2024).

Spatially-coordinated fluctuations and increase in correlative and recurrent activity from supra- to infragranular layers

In the horizontal dimensions, under higher correlation regimes, fluctuations are spatially coordinated and global within the central hexagon of the simulated subvolume (Figure 4H; Video 2). In depth, the size of fluctuations and level of correlated activity increases from supragranular to infragranular layers (Figure 4G1, I, left; Video 1), suggesting that activity in deeper layers of the model is more internally driven. We characterized the effect of recurrent connectivity by measuring the layer-specific decrease in firing rates when neurons in the model are disconnected from each other. As expected, we found that the effect (quantified by the ratio between connected and disconnected firing rates: RC/U) increased from supra- to infragranular layers (Figure 4I, right).

High-dimensional connectivity motifs predict layer-wise spontaneous activity

By analysing the underlying network structure, we observed a corresponding gradient in the topology of intra-layer synaptic connectivity from supra- to infragranular layers. We calculated the mean node participation of neuron populations in various dimensions (see Methods; Figure 4J, top). For dimension one, this is simply the degree of a neuron; for dimensions above one, this generalizes the notion to counting participation in dense, directed motifs of increasing size (directed simplices; Reimann et al., 2017). We found that correlations of this measure with the ratio of connected and disconnected firing rates increased with dimension (Figure 4J, bottom, 100% of data), indicating the importance of large directed simplices in shaping activity. Curiously, the correlations for higher dimensions were driven by a small number of neurons with very high node participation. This was indicated by a sharp drop in correlation when neurons above a given value were excluded (Figure 4J, bottom). Additionally, we studied the structural effect on the firing rate (here measured as the inverse of the inter-spike interval, ISI, which can be thought of as a proxy of non-zero firing rate). We found that for the connected circuit, the firing rate increases with simplex dimension; in contrast with the disconnected circuit, where this relationship remains flat (see Figure S6 red vs. blue curves and Methods).

This also demonstrates high variability between neurons, in line with biology, both structurally (Towlson et al., 2013; Nigam et al., 2016) and functionally (Wohrer et al., 2013; Buzsáki and Mizuseki, 2014). We next identified the cell types that are overexpressed in the group of neurons that have the 5% highest values of node participation across dimensions (Figure S7). This could inform theoretical point neuron models with cell-type specificity, for example. We found that while in dimension one (i.e., node degree) this consists mostly of inhibitory cells, in higher dimensions the cell types concentrate in layers 4, 5 and 6, especially for TPC neurons. This is in line with our structural layer-wise findings in Figure 8B in Reimann et al. (2024).

Stimulus-responses reproduce in vivo dynamics with millisecond-scale precision

Recreating simple whisker deflection experiments

We compared stimulus-responses with in vivo barrel cortex responses to both single whisker deflections and active whisker touches under anaesthetized (Reyes-Puerta et al., 2015b) and awake states (Yu et al., 2019) respectively. While the model is of non-barrel somatosensory regions, this nevertheless provides validations of overall excitability and the laminar structure of responses reflecting general trends of cortical processing. We activated a percentage (FP) of the VPM fibers projecting to the central column of the seven column subvolume (Figure 5A1). For each selected thalamic fiber, spike times were drawn from VPM PSTHs (peristimulus time histogram) recorded in vivo for the two stimulus types (Figure 5A2; Methods). For both stimulus types, a single thalamic stimulus was presented 10 times at 1 s intervals at three intensities (FP : 5%, 10%, 15%).

Layer-wise population responses to single whisker deflection and active whisker touch stimuli closely match in vivo millisecond dynamics and response amplitudes.

A1: Stimuli are simulated by activating a proportion (FP) of fibers projecting to the central hexagon of the seven column subvolumne. A2: Each fiber is assigned a spike time from a PSTH of VPM activity recorded in vivo for either a single whisker deflection or active whisker touch paradigm. B: Spiking activity and histograms (baseline and max normalised) for each layer-wise E and I population for the parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4, PF R = 0.3 and FP = 10% for a 2.5s section of the 10 whisker deflection test protocol. C: Trial-averaged PSTHs (baseline and max normalised) in response to the single whisker deflection stimulus for all of the parameter combinations which passed the initial criteria, and the in vivo references. D: Spatiotemporal evolution of the trial-averaged stimulus response in flat space for the same parameter combination in B. E: Left: Latencies of different layer-wise E and I populations for all criteria passing combinations in response to single whisker deflection stimuli. Right: Latencies for inhibitory interneuron subpopulations in response to active whisker touch stimuli. Correponding in vivo references are shown. F: Heatmap showing the effect of the meta-parameters on the difference of RE for the entire central column from the corresponding in vivo reference. Parameter combinations which failed the criteria tests are masked. G: Left: Ratio RE between maximum evoked response and spontaneous baseline activity by population. One line per simulation. Simulation line colour shows the difference of RE for the entire central column from the corresponding in vivo reference (color values shown in G). Thick black line shows median of in silico values. White shows in vivo reference. Right: Normalised ratios of each simulation by dividing by RE for the entire central column.

Validation of millisecond precise in vivo population responses and corresponding prediction of in vivo spontaneous firing rates

For the parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4, PF R = 0.3 and FP = 10%, E and I populations in each layer show clear responses on each trial (Figure 5B; three trials shown), except for L1. Videos 3 and 4 show all simulated combinations of the four meta-parameters. Out of 72 parameter combinations, 21 passed an initial assessment of in vivo similarity to the in vivo data based on latencies and decays (see Methods, Figure 5C, S5). Responses remained localized to within 600 µm of the stimulus location (Figure 5D, Video 5). In vivo, layer-wise I populations precede the corresponding E populations and response latencies increase with distance from thalamus, except for the fast response of L4 I. The latencies and decays are matched by criteria passing simulations (Figure 5C, E, left), except for a longer L4 I latency, and shorter and longer sustained responses of L2/3 E and L4 E. PV+ and Sst+ subpopulations respond reliably in layers 2-6, but the 5HT3aR+ subpopulation responses are weaker and less reliable (Figure S5). Under the active whisker touch paradigm, PV+ latencies precede Sst+ latencies in vivo, and either precede or are simultaneous with Sst+ latencies in silico (Figure 5E, right).

The passing parameter combinations form a contiguous region in the parameter space (Figure 5F), characterized by combinations of low PF R, [Ca2+]o and FP with higher ROU, which displayed weakly correlated spontaneous activity (compare Figure 5F with Figure 4G2). This offers a prediction that true in vivo firing rates are within the range of these lower FRs. Specifically, simulations with PF R from 0.1 to 0.5 robustly support realistic stimulus responses, with the middle of this range (0.3) corresponding with estimates of in vivo recording bias; both the previous estimates of Olshausen and Field (2006) and from a spike sorting study using this model (Laquitaine et al., 2024). The remaining results apply to the criteria passing simulations. We characterized the magnitude of responses by RE, the ratio between the peak of the evoked response and the pre-stimulus baseline activity level. Assuming that extracellular sampling biases are similar during spontaneous and evoked activity, RE provides a bias-free metric for comparing response magnitude with in vivo. Decreasing ROU and PF R, and increasing [Ca2+]o and FP led to a global increase of RE, and results matched in vivo closely for a large part of the parameter space (Figure 5F).

Interestingly, the parameter combination closest to in vivo was central in the parameter space of criteria passing simulations (Figure 5G, left). Values of RE, normalized by the mean RE across populations, were remarkably constant (Figure 5G, right), indicating that variation of the parameters provides a global scaling of response magnitude. More-over, I populations matched the in vivo reference closely, while the responses of L2/3 E were slightly greater, and for L4 E and L5 E slightly weaker. The region of the parameter space where responses match in vivo for the active whisker touch stimulus was the same as for single whisker deflections (Figure S8). Latencies were also similar to in vivo, with the main discrepancy being slower responses in L4 (E and I). With respect to the meta-parameters, we found that the best match for the anesthetized state was characterized by lower firing rates (PF R) and reduced external noise levels (lower ROU) compared to the awake active whisker touch paradigm (white areas in Figures 5F and S8B). These results fit expectation for anaesthetized and awake states respectively, and provide predictions of their spontaneous firing rates, noise levels and corresponding in silico regimes.

Prediction of 2-9% population response sparsity

Response sparsity (i.e. the proportion of neurons spiking at least once following a stimulus) varied across populations and meta-parameter combinations, but was around the range of 10-20% reported in vivo in the barrel cortex (Barth and Poulet (2012); Figure S5D). Interestingly, excitatory population responses were sparser than the corresponding inhibitory layer-wise populations across layers. The observed heterogeneous layer-wise population sparsities offer a prediction for in vivo. Moreover, the majority of responding cells spike only once to single whisker whisker deflections (Figure S5E), as reported in vivo (Isbister et al., 2021).

Validation and prediction through reproduction and extension of complex laboratory experiments using a single model parameterization

Building upon the in vivo-like activity states, we combined various simulation techniques to recreate laboratory experiments from the literature (see Methods). For these experiments, we use the parameter combination: [Ca2+]o = 1.05 mM, ROU = 0.4 and PF R = 0.3, as this combination supports realistic in vivo-like responses to the two stimulus types over a range of fiber percentages (FP).

Exploring the canonical model: layer-wise contributions to L2/3 stimulus-responses

In the canonical model of the cortex (reviewed e.g., in Lübke and Feldmeyer, 2007; Feldmeyer, 2012) information from the thalamus arrives to L4, then propagates to L2/3, from there to L5 (which serves as the main cortico-cortical output layer) and lastly to L6 (which projects back to the thalamus). The coordinated action of all layers of S1 is required for high-level behavioral tasks (Park et al., 2020). As the canonical model is based on the highest layer-wise density of axons, it cannot describe all interactions in the cortex. For example VPM does not only innervate L4, but also the border of L5 and L6, as well as the bottom of L2/3 (Meyer et al., 2010; Constantinople and Bruno, 2013; Sermet et al., 2019).

Recreating modulation of L2/3 PCs during optogenetic inactivation of L4 PCs during whisker deflection

To study how L4 contributes to the stimulus preference of L2/3 PCs, Varani et al. (2022) used optogenetic inactivation of L4 PCs during whisker stimulation and quantified the changes in the subthreshold response of L2/3 PCs. They found, that the early phase of the subthreshold response significantly differed from the control condition, if the whisker was deflected in either the most or the least preferred direction (see the top and bottom rows of their Figure 5B, C). From this they concluded that both L4 and VPM contribute to the direction tuning of L2/3 PCs. After reproducing their experimental conditions in silico (Figure 6A,B; Figure S9; see Methods) we confirmed that we can reproduce their results, i.e., subthreshold responses of L2/3 PCs decreased, when L4 PCs were inhibited (Figure 6C for preferred direction whisker stimulation; see Methods).

Reproducing and extending detailed experiments exploring the canonical model and encoding of contrast, rate-coded and synchronous information.

First experiment: Reproducing the effect of optogenetic inactivation of L4 pyramidal cells on L2/3 stimulus responses and predicting the role of other pathways through circuit lesions. A: Schematics of whisker kinetics and VPM fiber rates during 500 ms long whisker hold stimulus. Fraction of VPM fibers coding for each kinetic feature are taken from Petersen et al. (2008). B: Mimicking the effect of activation of the Halo inhibitory opsin in silico. Injected somatic hyperpolarizing current mimicking opsin activation (top), and the resulting somatic voltage trace from a combination of injected conductance, current, and synaptic PSPs from the network (bottom). C: Comparison of average traces from selected L2/3 PCs (see Methods) in control (black) and optogenetically inhibited (green) conditions. D: Same as C, but instead of mimicking the optogenetic inhibition of L4 PCs, only the connections to L2/3 PCs are “cut” (compare inset with the one in C). The right part depicts connections systematically cut from PCs in all layers, while the left shows L4 only for a better visual comparison with the conditions of Varani et al. (2022) in C.

Second experiment: Contrast tuning modulation by optogenetic activation of PV+ interneurons. E: Spatial distribution of firing rates of VPM fibers projecting to a single simulated column at different points in time, corresponding to linear drifting gratings with a temporal frequency of ftemp = 2 Hz and a spatial frequency of fspat = 0.001 cycles/µm. F: Firing rate signal of a single VPM fiber corresponding to a random sequence of drifting grating patterns with different contrast levels C, as indicated by the legend. All optogenetic stimuli targeting PV+ (or Sst+) interneurons were completely overlapping with the grating stimulus intervals, as indicated by the blue bars. G1: Spiking activity and PSTH of an exemplary PV+ interneuron over 10 repetitions (trials) in response to a grating stimulus at maximum contrast (C = 1.0) without (control) and with medium strength (150%) optogenetic stimululation. H1: Contrast tuning modulation of an exemplary PV+ interneuron by different levels of PV+ optogenetic stimulation, ranging from 0% (control) to 300%. The individual markers denote actual data points (mean SD over 10 repetitions, normalized to baseline response at maximum contrast) while curves illustrate sigmoidal fits to these data points. G2/H2: Same as C1/D1, but for exemplary PCs. H2 inset: Illustration of sigmoidal parameters m, n, Rmax, and c50.

Third experiment: Encoding of synchronous and rate-coded signals by interneuron subtypes. I: A binary signal is encoded either through changes in the rate or synchronicity of optogenetic pulses applied to a set of PCs. We define a direct correspondence between single optogenetic pulses and single spikes. Spiking activity is measured in PV+, Sst+ and 5HT3aR+ cells, and the mutual information between the binned activity of individual cells and the original binary signal is measured. J: Visualization of the rate (upper) and synchronous (lower) coded stimulus experiments stimulating 1000 PCs, showing the binary signal (top), input neuron spike trains for 40 neurons (middle), and responses of the three L2/3 neuron types (bottom). K: Upper: Results for the rate coded stimulus experiment. Left: Mutual information between spiking activity and the binary signal (one point for each cell that spiked at least once). Only activity in the 50 ms following the change of the binary signal is considered. Cells with mutual information not significantly different from that of a shuffled control are coloured grey. Centre: Same as left but considering all time bins. Right: The effect of the number of stimulus neurons on the mean single cell mutual information for each subpopulation. Lower: The same as upper but for the synchronous stimulus type.

Prediction that L4 PCs enhance L2/3 PC responses directly, not through disynaptic inhibition

We then leveraged our in silico setup to study what Varani et al. (2022) could not, because of methodological limitations. In our reading, the authors aimed to test how direct E connections from L4 PCs to L2/3 PCs influence the stimulus representation in L2/3. This connection can not specifically be blocked in vivo, instead (95% of) the L4 PC population is inhibited (as well as some lower L3 PCs). In our setup we could selectively block the connection and found almost the same result (compare Figure 6C and D, left). This extends the conclusion of Varani et al. (2022): L4 PCs contribute to the stimulus preference of L2/3 PCs via direct E connections, and not via disynaptic inhibition.

Prediction of the contribution of different layer-wise populations to L2/3 PC response

The authors also discussed studying L5 PCs’ contribution to L2/3 responses (as a large fraction of L5 PC axons terminate in L2/3), but this is infeasible with current mouse lines. Leveraging our model, we found that L5 contributes much less to subthreshold L2/3 traces than L4 (Figure 6D, right). Extending to other presynaptic layers, we found that the contribution of L2/3 is similar to that of L4, whereas inputs from L6 are negligible (Figure 6D, right). Whilst mouse lines targeting L5 PCs might arrive soon (which could validate our prediction), blocking L2/3 connections between L2/3 cells without hyperpolarizing the same L2/3 population seems only achievable in silico.

Recreation of contrast tuning modulation by optogenetic activation of PV+ interneurons

Shapiro et al. (2022) compared the modulatory effects of optogenetic activation of interneuron subtypes on contrast tuning of neurons in V1. Whilst visual regions differ from somatosensory regions, the study provided an opportunity to explore the generality of interactions between E and I subpopulations. We therefore created a version of the study in silico, presenting spatio-temporally modulated patterns with various contrast levels through the VPM inputs (Figure 6E,F; see Methods).

Overall, we found within the single simulated column 228 PCs and 259 PV+ interneurons with robust contrast tuning of their responses (see Methods). Within spike detection range of a vertically penetrating extracellular electrode (50 µm; Henze et al. (2000); Buzsáki (2004)), we found 15 PCs and 13 PV+ interneurons with robust tuning, which is roughly comparable to the original study (average of 8.2 tuned PCs and 3.0 PV+ neurons per mouse Shapiro et al., 2022). Unlike the original study we found no Sst+ neurons with robust tuning.

Additionally, we found the following results, all in line with (Shapiro et al., 2022): The firing rates of both PCs and PV+ neurons were affected by optogenetic activation of the PV+ subpopulation (Figure 6G), resulting in changes to the contrast tuning curves of these populations (Figure 6H, S12A). The main effect of the optogenetic manipulation was an increase in contrast tuning for PV+ and a decrease for PCs. However, at high contrasts, we observed a paradoxical effect of the optogenetic stimulation on L6 PV+ neurons, reducing their activity with increasing stimulation strength (Figure S12B; cf. Mahrach et al. (2020)). This effect did not occur under grey screen conditions (i.e., at contrast 0.0) with a constant background firing rate of 0.2 Hz or 5 Hz respectively (not shown). The individual tuning curves could be accurately fit by sigmoidal functions and consequently the effect of optogenetic activation could be quantified in terms of changes to the fitted parameters (Figure 6H2 inset, S13; Methods). We found great variability between neurons with increasing trends for parameter m in PV+ and c50 in PCs, and decreasing trends for Rmax in PV+ and both m and Rmax in PCs. Unlike in (Shapiro et al., 2022), we found in addition an increasing trend of c50 also in PV+ interneurons (Figure S13).

Comparing three abstract mathematical models (Shapiro et al., 2022; Methods), we found that the saturation additive model best captured the effect of the optogenetic manipulation for PV+ neurons (Figure S14A, B1), as in the reference study. The relative contributions of saturation and addition varied strongly between individual neurons (Figure S14C1). At low intensities of the optogenetic manipulation, this distribution matched qualitatively the results of Shapiro et al. (2022), but at higher intensities we observed a shift towards stronger saturation. Combining the saturation additive model with a simplified, population-level model of PC firing (Shapiro et al., 2022; Methods) provided a good description of the observations (Figure S14B2) with relative contributions of saturation and addition that qualitatively match the reference (Figure S14C2).

Encoding of synchronous and rate-coded signals by inhibitory subpopulations

Prince et al. (2021) studied how different I subpopulations represent signals that are encoded as changes in the firing rate or synchronicity of PCs (Figure 6I). We repeated their study in silico (Figure 6J), but using in vivo-like conditions as the authors have suggested in their Discussion. The original study pointed out the importance of membrane properties and short-term dynamics of synaptic inputs onto the subpopulations in shaping the results. Note, that the lower synaptic reliability in vivo weakens synaptic depression (Borst, 2010), and the ongoing background input leads to higher membrane conductance (Destexhe et al., 2001); therefore our results are expected to differ substantially, and provide an independent prediction for a different dynamic state. While in the original in vitro study stimulation of 10 PCs was sufficient to find a decodable signal, we expected a larger number to be required under our in vivo-like conditions, due to the lower [Ca2+]o, and the higher level of noise (see Methods).

Indeed, we found little discernible mutual information in I spiking for the activation of 10 PCs, but that mutual information increased with the number of activated neurons (Figure 6K, right). For 1000 activated neurons each of the three I subpopulations in L2/3 showed clear but qualitatively unique modulation for both encoding schemes (Figure 6J). As in (Prince et al., 2021) we found differences in the encoding capabilities of the I subpopulations (Figure 6K). We found strongest encoding in PV+ neurons for both schemes, and overall low encoding strength in Sst+ neurons. Further, we found overall low amounts of mutual information for the synchronous stimulus, which is in line with Prince et al. (2021), as they linked encoding of that stimulus to depressing synaptic dynamics, which would be weakened in vivo.

Exploring the effect of inhibitory targeting trends observed in electron microscopy (EM)

Additional targeting rules recreate quantitative EM-connectivity trends

We next explored how certain trends in inhibitory targeting observed in a recent mouse electron microscopic connectome (Schneider-Mizell et al., 2023; Figure 7A) affect spontaneous and evoked activity. Consequently, an alternative connectome which quantitatively recreates these trends has been built (Reimann et al., 2024). We refer to this as the Schneider-Mizell compatible connectome (SM-connectome). It ensures that PV+ neurons target the perisomatic region, VIP+ neurons target other inhibitory neurons, and cells in L1 connect predominantly monosynaptically. Consequently, inhibition onto E neurons is sparser and closer to the soma (Figure 7A2). After validating the presence of these trends, we characterized how this affects the dynamics of the spontaneous and evoked states (Figure 7A3). Note, that individual synapses were recalibrated to preserve the pathway specific reference PSP amplitudes. As expected, the largest change was for PV+ neurons where the average number of vesicles in the release-ready pool NRRP was increased by 2.4 times.

The effect of inhibitory targeting trends observed in electron microscopy on spontaneous and stimulus-evoked activity.

A1: Panel illustrating interneuron synaptic targeting features and caption taken from Schneider-Mizell et al. (2023) (under CC-BY 4.0 license). i) For determining inhibitory cell subclasses, connectivity properties were used such as an axon (green) making synapses (green dots) the perisomatic region of a target pyramidal cell (purple). ii) Dendritic compartment definitions for excitatory neurons. iii) Cartoon of a multisynaptic connection (left) and the synapses within the multisynaptic connection considered “clumped” along the presynaptic axon (right). A2: Exemplar neurons showing the synapse locations from different inhibitory m-type groupings in the original and SM connectome respectively: Distal targeting (DistTC: martinotti, bipolar, double bouquet cells), Perisomatic targeting (PeriTC: large basket, nest basket cells), Sparse targeting (SparTC: descending axon, neurogliaform, horitzontal axon cells), Inhibitory targeting (InhTC: bitufted, small basket cells). A3: Heatmaps showing the mean fraction of efferent synapses from the inhibitory m-type groupings onto inhibitory neurons, somas, proximal dendrites, apical dendrites, distal dendrites, or that are part of clumped or multisynaptic connections. Heatmaps show mean fractions (left to right) for the original, Schneider-Mizell et al. (2023) characterization of the MICrONS dataset ((MICrONS-Consortium et al., 2021)), SM-connectome, and the change between the original and SM-connectome. B1: Proportional change in mean conductance injection between original and SM-connectome by population. B2: Distribution of ratios between connected and disconnected firing rates (RC/U) by population over non-bursting meta-parameter combinations for original and SM-connectome. B3: Comparison of correlation (r-values) between E and I histograms (5 ms bin size, 1σ Gaussian smoothing; central column) between original and SM-connectome (each point represents a specific meta-parameter combination). C: Scatter plot showing the change in the mean number of afferent inhibitory synapses onto each E (red) and I (blue) population after calibration versus the proportional change in mean conductance injection. The number of sides of the shape indicates the layer (triangle used for L2/3). D1: Proportional change in the the ratio between the peak of the evoked response and the pre-stimulus baseline activity level (RE) between the original and SM-connectome. Plot shows distribution over meta-parameter combinations which passed the the initial assessment of in vivo similarity to the in vivo data based on latencies and decays (see Methods). D2: Change in the latencies of populations peak responses between original and SM-connectome (distribution over same meta-parameter combinations shown in D1). E: Mean membrane potential responses for layer-wise subpopulations for the meta-parameter combination [Ca2+]o = 1.05 mM, ROU = 0.4, PF R = 0.3 and FP = 10%. Responses averaged over 20 stimulus repetitions with an inter-trialinterval of 500ms.

EM targeting trends increase inhibition of inhibitory populations enabling increased long-range excitation

Inspite of this, the change in connectivity affected the activity sufficiently that recalibration of missing synaptic compensation was needed to attain the target firing rates for the original meta-parameter combinations (Figure S15). Mean conductances onto I neurons had to be increased, indicating the efficacy of the newly introduced I-I targeting by VIP+ neurons (Figure 7B1). On the other hand, conductances for E neurons remained roughly the same, except for L6. Curiously, we found that the firing rates of I populations were reduced when the internal connectivity was activated compared to when it was inactive, further highlighting the efficacy of the I-I targeting (Figure 7B2, S15). For E neurons in L2-5, activating internal connectivity increased firing rates, but less than in the original connectome. As the number of inhibitory synapses innervating them decreased (Figure 7A2), this indicates that the placement of PV+ synapses closer to the soma increased the efficacy of inhibitory as expected. Taken together, this indicates that in the SM-connectome, I populations are more externally driven and more internally inhibited (as a function of proximity to L4), and L6 E more internally driven during spontaneous activity. Over the different meta-parameter combinations, correlation in the network decreased relative to the original circuit with some exceptions (Figure 7B3, S15). This indicates, that activity in the SM-connectome is shifted slightly towards a more asynchronous state.

By correlating the structural and functional changes we concluded that the increased conductance injection into the I populations can be explained by the increased inhibitory synapse count (Figure 7C). Conversely, the mostly decreased synapse counts for E populations did not lead to a decrease of conductance injection, further indicating that the more perisomatic positioning of PV+ synapses compensates for the reduced synapse counts.

EM targeting trends affect population-wise stimulus-evoked response magnitudes and latencies

The lack of effect onto most of the E populations disappeared in the evoked state (Figure 7D). Overall, responses increase strongly in L23 and slightly decreased in L5 and L6 (Figure 7D1), while latencies of responses increased in every layer (Figure 7D2), suggesting different roles of the more specific inhibitory targeting in different layers.

EM targeting trends hyperpolarize Sst+ and HT3aR+ late response, and disinhibit L5/6 E

Studying somatic membrane potentials for different subpopulations in response to whisker deflections shows that PV+, L23E and L4E subpopulations are largely unaffected in the SM-connectome (Figure 7E). Interestingly, Sst+ and 5HT3aR+ subpopulations show a strong hyperpolarization in the late response that isn’t present in the original connectome. Interestingly, this corresponds with a stronger late response in L5/6 E populations, which could be caused by disinhibition due to the Sst+ and 5HT3aR+ hyperpolarization. This could be explored further in follow up studies using our connectome manipulator tool (Pokorny et al., 2024).

Mid-range connectivity determines independent functional units and selective propagation of stimulus-responses at larger spatial scales

Stable spontaneous activity only emerges in nbS1 at predicted in vivo firing rates

After calibrating the model of extrinsic synaptic input for the seven column subvolume, we tested to what degree the calibration generalizes to the entire nbS1. Notably, this included the addition of mid-range connectivity (Reimann et al., 2024). The total number of local and mid-range synapses in the model was 9138 billion and 4075 billion, i.e., on average full model simulations increased the number of intrinsic synapses onto a neuron by 45%. Particularly, we ran simulations for PF R [0.1, 0.15, …, 0.3] using the OU paramaters calibrated for the seven column subvolume for [Ca2+]o = 1.05 mM and ROU = 0.4. Each of these full nbS1 simulations produced stable non-bursting activity (Figure 8A), except for the simulation for PF R = 0.3, which produced network-wide bursting activity (Video 6). Activity levels in the simulations of spontaneous activity were heterogeneous (Figure 8B, Video 7). In some areas, firing rates were equal to the target PF R, whilst in others they increased above the target (Figure 8C). In the more active regions, mean firing rates (averaged over layers) were on the order of 30-35% of the in vivo references for the maximum non-bursting PF R simulation (target PF R : 0.25). This range of firing rates again fits with the estimate of firing rate bias from our paper studying spike sorting bias (Laquitaine et al., 2024) and the meta-parameter range supporting realistic stimulus responses in the seven column subvolume. This also predicts that the nbS1 cannot sustain higher firing rates without entering a bursting regime.

Full nbS1 simulations: independent functional units, millisecond-precise stimulus-responses, selective propagation of stimulus-evoked activity through mid-range connectivity, linking structure to function.

A: Spiking activity and max normalised histograms for layer-wise E and I populations for the meta-parameter combination [Ca2+]o: 1.05 mM, ROU : 0.4, PF R: 0.15, and the 0th (upper) and 39th (lower) hexagonal subvolumes (shown in B). B1: Mean FRs across the model. Columns 0 and 39 (subsequently used) are highlighted. B2: Spiking correlations of E/I populations in 240 hexagonal subvolumes (diameter: 460 µm). C: Target vs. observed PF R values for the four non-bursting simulations. The number of sides of each marker corresponds to the layer (triangle represents L2/3). Red: E, blue: I.. D: Correlations between E and I populations by layer for the four non-bursting simulations for the two hexagons (top and bottom). Color of line from light to dark represents PF R. Middle: Distributions of spiking correlations of layer-specific populations in columns (520 µm) of the nbS1 model (black dots; mean values: grey). Compared to the central column when the seven column subvolume is simulated in isolation (green). E: Trial-averaged activity in flatspace following single whisker deflections. Lower: Subset of time windows shown with only the top 2% of bins for each time window after baseline activity level subtracted. F: Trial-averaged PSTHs (baseline and max normalised) for column 0. For FP = 15% and FP = 25% respectively, the evoked responses passed and failed the latency and decay-based criteria tests for similarity to in vivo. Response for FP = 20% passed the tests for all populations except L5 E which showed a more sustained response. G: Trial-averaged response of column 39. H: Correlations of mid-range inputs. Upper: Distribution of spiking correlations of inputs into 240 subvolumes. Calculated at reduced spatial resolution, based on connection counts and correlations between the subvolumes (see Methods). Subvolumes are sorted by the E/I correlation of neurons within them. Lower: Mean of the correlations of inputs into 240 subvolumes vs internal spiking correlation for all subvolumes during spontaneous activity. Black line: linear fit. Calculated based on connection counts and correlations between 50 µm hexagonal subvolumes (see Methods). Data for evoked activity is shown in Figure S16. I: Upper: Spiking correlations of subvolumes against their distances in spontaneous (red) and evoked (orange: Fp = 0.15%, blue: Fp = 0.2%, green: Fp = 0.25) activity. For increased spatial resolution, smaller (58 µm) subvolumes were used, hence correlation values are not comparable to values in B. Lower: Distribution of soma distances of neuron pairs connected by local (orange) or mid-range (blue) connectivity. Green: distances of all pairs, independent of connectivity. J: Upper: Classification of subvolumes based on low or high internal correlation and membership in mid-range rich club Data for evoked activity shown in Figure S16. Lower: E/I correlation of subvolumes for members of the rich club (red) and non-members (blue). K: FRs of subvolumes following a stimulus with FP = 0.15% (top) and FP = 0.2% (bottom). Mean (lines) and SEM (shaded area) over 10 repetitions shown for subvolumes directly innervated by the VPM stimulus (black), strongly innervated by directly innervated subvolumes (with over 106 synapses, red), or by medium strength (over 105 synapses, yellow) or weak (over 104 synapses) indirect innervation. Dashed lines indicate locations of peaks.

Independent functional subunits emerge during spontaneous activity with accurate layer and distance-dependence of noise correlations

The subregion with consistently higher firing rates was the upper-limb representation (S1ULp), separated from subregions of lower activity by the dysgranular zones (S1DZ, S1DZO). This demonstrates that the larger spatial scale of the model supports several independently acting functional units. Locations with higher firing rates were accompanied by higher spiking correlations of the local population (Figure 8B2, Video 7), indicating activity that was more driven by intrinsic excitation. Unlike in the simulations of smaller volumes, the correlation emerged mostly from short, transient packets of activity in L6 (Figure 8A). Consequently, deviation from the target PF R in those cases was largest in L6 (Figure 8C). Taken together, it appears as if the additional synapses from mid-range connectivity moved the model along the spectrum from extrinsically vs. intrinsically driven, but further for some regions than for others. We confirmed a divergence into two populations: one with correlation slightly increased from the simulations of smaller volumes, and one with almost maximal correlations (Figure 8D). This distinction was determined mostly in L6 with increases of correlations in other layers remaining small.

The distance dependence of correlations followed a similar profile to that observed in a dataset characterizing spontaneous activity in the somatosensory cortex (Reyes-Puerta et al., 2015a) (compare red line in Figure 8I with Figure S16). In the in vivo dataset spiking correlation was also low but highest in lower layers, with short “up-states” in spiking activity constrained to L5 & 6 (see Figure 1E,F in (Reyes-Puerta et al., 2015a)). In the model, they are constrained to L6.

Stable propagation of evoked activity through mid-range connectivity only emerges in nbS1 at predicted in vivo firing rates

We repeated the previous single whisker deflection evoked activity experiment in the full model, providing a synchronous thalamic input into the forelimb sub-region (S1FL; Figure 8E; Video 8 & 9). Responses in S1FL were remarkably similar to the ones in the seven column subvolume, including the delays and decays of activity (Figure 8F). However, in addition to a localized primary response in S1FL within 350 µm of the stimulus, we found several secondary responses at distal locations (Figure 8E; Video 9), which was suggestive of selective propagation of the stimulus-evoked signal to downstream areas efferently connected by mid-range connectivity. The response of the main activated downstream area (visible in Figure 8E) was confined to L6 (Figure 8G). In a follow up study using the model to explore the propagation of activity between cortical regions (Bolaños-Puchet and Reimann, 2024), it is described how the model contains both a feedforward projection pattern, which projects to principally to synapses in L1 & L23, and a feedback type pattern, which principally projects to synapses in L1 & L6. On visualizing the innervation profile from the stimulated hexagon to the downstream hexagon we can see that we have stimulated a feedback pathway (Figure S16).

Membership in mid-range connectivity rich club predicts correlation of spontaneous activity

As the model is homogeneous across the different subregions in terms of neuronal morphologies and physiology, we expected the differences in spontaneous and evoked activity across the model to be explained by the topology of synaptic connectivity (Reimann et al., 2024). While the extrinsic input from the OU injections was uncorrelated by design, synaptic input from local connectivity would feature at least some degree of correlation. This led to asking, how correlated was the input from mid-range connections? We estimated for each pair of mid-range connections innervating a 460 µm hexagonal subvolume its spiking correlation and compared the resulting distribution to the correlation within the subvolume itself (see Methods; Figure 8H, upper). We found low correlations for most pairs, but with a significant number of high-correlation outliers for subvolumes with high internal correlations. Moreover, when we considered the mean of the measure, we found that it explains most of the variability in internal correlations, for both spontaneous (Figure 8H, lower) and evoked states (Figure S16).

This led to the following tentative explanation: Local connectivity provides inputs from within 500 µm, which is virtually guaranteed to be correlated to some degree (Figure 8I upper vs. lower, orange). This is a result of connections being locally dense, as they are concentrated within a small neighborhood (Figure 8 lower, green vs. orange). mid-range connectivity originates from further away, sampling from different functional units that may or may not be correlated, and being more diluted (Figure 8I lower, green vs. blue). Consequently, mid-range inputs are likely to act not much different from the uncorrelated OU process, unless this is overcome by a non-random topology of mid-range connectivity.

Specifically, we considered the mid-range connectivity at meso-scale in terms of the graph of synapse counts between 460 µm subvolumes. When the graph was thresholded at eight million synapses, most subvolumes lacked connections at that strength completely, but others formed a single connected component, indicating the formation of a rich club (Colizza et al., 2006; Figure 8J, upper, red and orange). We found that subvolumes with high internal correlation were part of the rich club, or directly adjacent to one of its members (Figure 8J). In the presence of strong thalamic inputs, this could be overcome, as the increased activity level spread along sufficiently strong mid-range connections (Figure 8K), albeit with substantial delays between 60 and 100 ms. This could induce high correlations in strongly connected parts of the model, outside the rich club as well (Figure S16) and shows that the model supports selective propagation of activity to areas efferently connected with mid-range connectivity.

Discussion

We presented novel methods to build, simulate and validate models of cortical tissue that correspond directly with 3D digitised brain atlases. As demonstrated, this enables laboratory experiments to have a simulatable in silico counterpart, and vice versa: predictions made by the model automatically have a precise correspondence in biology. By recreating and extending five laboratory experiments under a single model parameterization, we provided strong model validation and demonstrated the model’s natural versatility. To our knowledge the simulations of the full nbS1 represent the first simulations of stable in vivo-like spontaneous and stimulus-evoked activity in a large-scale biophysically detailed model of multiple cortical subregions connected through local and mid-range connectivity. The model generated an initial set of predictions about the relationship between cortical structure and function, and provides a basis for future studies and predictions. Table S1 describes each quantitative prediction made in this manuscript, and how it might be tested in vivo. Additionally, all assumptions are listed in Table S8. We made the model and simulation tools available for use, enabling community driven testing, exploration and refinement.

Excitation from non-modeled brain regions is modeled by injecting somatic conductances and is calibrated for specific populations. For the purpose of calibration, we presented a novel technique, which allows rapid and methodical characterisation and validation of emergent model activity. The efficiency of the technique allows rapid recalibration after changes to the model such as adding new brain regions or changing anatomical details. Connected divided by unconnected firing rates measured during the process are also a prediction of the degree to which the activity of a population is determined by incoming connections from local vs extrinsic sources. We found an increase of the importance of local connectivity from superficial to deeper layers, in line with the canonical view of the cortex (Lübke and Feldmeyer, 2007; Feldmeyer, 2012), which places layers 5 and 6 at the end of local information flow.

Although abstract relative to the model, the injection amplitudes could be predicted from the number of missing synapses, suggesting it provides a functional representation of their anatomical counterparts. The modular model of extrinsic connectivity can be refined or replaced in the future, or when larger fractions of the brain are modeled. The calibration framework could optimize per population parameters for other compensation methods, whilst still offering an interpretable spectrum of firing rate regimes at different levels of PF R. For example, more realistic compensation schemes could be explored which introduce a) correlations between the inputs received by different neurons and b) compensation distributed across dendrites, as well as at the soma. We predict that such changes would make spontaneous activity more correlated at the lower spontaneous firing rates which supported in vivo like responses (PF R : 0.1 − 0.5), which would in turn make stimulus-responses more noise correlated.

Our approach to obtaining in vivo-like activity contrasts with a recent approach in a large-scale biophysical model of primary visual cortex of similar scale to our smaller subvolume (Billeh et al., 2020). There, extrinsic inputs were delivered through dendritic synapses instead of somatic injection, which is more anatomically realistic, and intrinsic recurrent weights were adjusted to match extracellularly recorded firing rates. They find a single activity state, instead of the in vivo-compatible spectrum here that allowed us to contrast anesthetized and awake states. The model in general differs from hybrid models, which jointly use biologically-detailed models and point neuron models (Billeh et al., 2020; Dura-Bernal et al., 2023), and also from the work of Egger et al. (2020) who modeled activation of L5 pyramidal cells by constraining patterns of synaptic input based on receptive fields and predicted anatomical innervation.

Given the complexity of the model, how can we be confident that its activity is in vivo-like? First, we demonstrated correlated spontaneous dynamics in the form of global and local fluctuations and dynamic E-I balance, as found experimentally (Renart et al., 2010). Second, population firing rate distributions were long-tailed with sub 1 Hz peaks, and were similar for spontaneous and stimulus-evoked activity, as reported (Wohrer et al., 2013). Moreover, mean firing rates below 1 Hz are required by metabolic constraints (Attwell and Laughlin, 2001; Lennie, 2003). In response to simple whisker stimuli, response sparsity, spike counts, and the temporal profile and amplitudes of layer-wise populations were similar to in vivo under anesthetized and awake conditions, for different regions of the meta-parameter space. The effect of the spontaneous meta-parameters on evoked responses, also shows that the networks spontaneous state affects stimulus-responses, as observed in vivo (Isbister et al., 2021). Although from the barrel system, these stimuli offer some of the simplest stimulus-response paradigms, enabling principles of neural dynamics and information processing to be studied; in particular, the correspondence between atomic units of sensory information and neural representations. The model also predicts the number of thalamic fibers stimulated for whisker-flick stimuli. Such validations are crucial for complex models, as they provide context for more complex validations. Moreover, in a rich and complex nonlinear system, the source of discrepancies with in vivo activity are simpler to ascertain under simpler protocols.

For more complex validations, we reproduced protocols from three laboratory experiments. These experiments used a wide array of techniques, which could be accurately recreated due to the model’s close correspondence with biological tissue. Additionally, we could go beyond the original experiments. For example, we predicted an increased number of neurons would be required to have a decodable effect on postsynaptic activity under in vivo conditions over the in vitro conditions of the original experiment. This can be explained by lower synaptic reliability under in vivo conditions. Also, we predicted a paradoxical effect of optogenetic stimulation on L6 PV+ interneurons, namely a decrease in firing with increased stimulus strength. This is reminiscent of the paradoxical responses found by Mahrach et al. (2020) in the mouse anterior lateral motor cortex (in L5, but not in L2/3) and barrel cortex (no layer distinction) respectively. While Mahrach et al. (2020) conducted their recordings in awake mice not engaged in any behavior, we observed this effect only when drifting grating patterns with high contrast were presented. Nevertheless, consistent with their findings, we found the effect only in deep but not in superficial layers, and only for PV+ interneurons but not for PCs. Our model could therefore be used to improve the understanding of this paradoxical effect in follow up studies. These examples demonstrate that the approach of modeling entire brain regions can be used to further probe the topics of the original articles and cortical processing. This could be continued in the future using paradigms that take advantage of the multiple regions offered by the model to study processing in cortical hierarchies.

Responses to whisker flick stimuli were closer to biology for values of the PF R meta-parameter between 0.1 and 0.5, according with the known inflation of mean firing rates stemming from the bias of extracellular spike sorting techniques towards larger and more active neurons (Olshausen and Field, 2006). Supporting our estimates, patch-clamp experiments show ubiquitous neuronal silence (Crochet and Petersen, 2006), and spontaneous and stimulus-evoked barrel cortex firing rates as much as 10 times lower than for extracellular recordings (0.05-0.15 Hz vs. 0.8-1.5 Hz; Olshausen and Field, 2006). This is also inline with our recent work using the model, which estimated a spike sorting bias corresponding to PF R = 0.3 using virtual extracellular electrodes (Laquitaine et al., 2024). Each recording technique comes with challenges and biases (Shoham et al., 2006; Olshausen and Field, 2006; Barth and Poulet, 2012; Wohrer et al., 2013). Calcium imaging and patch-clamp experiments are less biased towards frequently spiking neurons (Wohrer et al., 2013), yet the former can only infer spiking activity and may be biased towards neurons that favor marker expression. Patch-clamp techniques can reduce firing rates, be biased towards larger or more active neurons (Olshausen and Field, 2006); and are limited to recording a few neurons simultaneously, preventing characterizations of single trial population dynamics. An earlier version of the model has been used to simulate cortical extracellular potentials (Reimann et al., 2013). We can now simulate modern extracellular electrodes and compare multi-unit activity and spike sorting results to the in vivo-like ground truth to gain further insights into the potential over-estimation of mean firing rates.

During simulations of the entire nbS1, we found emerging spatial inhomogeneities in network activity. Specifically, we observed sharp transitions in firing rates and pairwise correlations at disgranular zones. This highlights that the model supports several, independently acting functional units. Analysis of spatial correlations suggests a radius of approximately 400 µm for a single unit. This property was not present in the previous intra-regional scale model (Markram et al., 2015), but is rather an emergent property of the interregional scale, demonstrating that we achieved our goal of building a model that can be used to study mid-range interactions between regions. Analyzing the source of the inhomogeneities suggested a prominent role of a non-random rich-club structure of mid-range connections at the meso-scale. Within functional units, we found that inhomogeneous increases of spiking correlations were confined to spontaneous packets of L6 activity.

In keeping with the philosophy of compartmentalization of parameters and continuous model refinement (see Introduction), it was essential to improve validity at the columnar scale (relative to Markram et al. (2015)) as part of demonstrating validity of the full nbS1. Indeed, improved parametrization and validation at smaller scales was essential to parameterizing background input which generated robust nbS1 activity within realistic [Ca2+]o and firing rate ranges. We view this as a major achievement, as it was unknown whether the model would achieve a stable and meaningful regime at the start of our investigation. Whilst we would have liked to go further, our primary goal was to publish a well characterized model as an open resource that others could use to undertake further in-depth studies. In this regard, we are pleased that the parametrization of the nbS1 model has already been used to study EEG signals (Tharayil et al., 2024), as well as propagation of activity between two subregions (Bolaños-Puchet and Reimann, 2024). Investigation, improvement and validation must be continued at all spatial scales in follow up papers with detailed description, figures and analysis, which cannot be covered in this manuscript. Each new study increases the scope and validity of future investigations. In this way, this model and paper act as a stepping stone towards more complex questions of interest to the community such as perception, task performance, predictive coding and dendritic processing. This was similar for Markram et al. (2015) where the initial paper was followed by more detailed studies. Unlike the Markram et al. (2015) model, the new model can also be exploited by the community and has already been used in a number of follow up papers studying (Ecker et al., 2024a,b; Bolaños-Puchet and Reimann, 2024; Pokorny et al., 2024; Egas Santander et al., 2024; Tharayil et al., 2024; Laquitaine et al., 2024). We believe that the number of use cases for such a general model is vast, and is made larger by the increased size of the model.

It is important to highlight where the model deviates from biological data. First, the model omits several known anatomical elements, such as glia and the presence of gap junctions between certain neuron-types, but the model’s spatial context provides a natural scaffold for them. Our modeling techniques assume generality of rules and parameters, unless indicated by the data. Therefore, violations indicate that biological rules and data may be more specific. For example, model thalamic inputs innervate any dendrite placed in the targeted layers equally, but this did not reproduce the timing of inhibitory subpopulation responses. Stronger innervation of PV+ neurons than other inhibitory neurons may rectify this. However, it is unclear whether innervation should be anatomically or physiologically stronger, as the best available data mixes structural and physiological strength (Sermet et al., 2019).

In our previous work (Roussel et al., 2023) we linked mouse inhibitory me-models to transcriptomic types (t-types) in a whole mouse cortex transcriptomic dataset (Gouwens et al., 2019). This can provide a direct correspondence in future large-scale mouse models. As we model only a single electrical type for pyramidal cells there is no one-to-one correspondence between our me-models and the 10 different pyramidal cell types identified there. We are not currently aware of any method which can recreate the electrical features of different types of pyramidal cells using only generic ion channel models. To achieve the firing pattern behavior of more specific electrical types, usually ion channel kinetics are tweaked, and this would violate the compartmentalization of parameters. In future we hope to build morpho-electric-transcriptomic type (met-type) models by selecting gene-specific ion channel models (Ranjan et al., 2019, 2024) based on the met-type’s gene expression. Data specific to different neuron sections (i.e. soma, AIS, apical/basel dendrites) of different met-types, such as gene expression, distribution of ion channels, and voltage recordings under standard single cell protocols would be particularly useful.

Finally, when required, we generalized data from different animals and brain regions to build our rat nbS1 model. This is the accepted state-of-the-art in computational neuroscience, e.g., all 19 data-driven models of rodent microcircuitry listed in Figure 2 of the recent review of Ramaswamy (2024) conduct some sort of mixing, including the advanced mouse V1 model of the Allen Institute (Billeh et al., 2020). Whilst a first truly single species model would be a great advance, it is not required to be the basis of in silico research. Our model (and previous ones) can be used to study the many cortical mechanisms that are common to the closely related organisms rat and mouse. This paper and its companion paper serve to present a methodology for modeling micro- and mesoscale anatomy and physiology, which can be applied for other cortical regions and species. With the rapid increase in openly available data, efforts are already in progress to build models of mouse brain regions with reduced reliance on data mixing thanks to much larger quantities of available atlas-based data. This also includes data for the validation of emergent network level activity. Here we chose to compare network-level activity to data mostly from the barrel cortex, as well as a single study from primary visual cortex. Whilst a lot of the data used to build the model was from the barrel cortex, the barrel cortex also represents a very well characterized model of cortical processing for simple and controlled sensory stimuli. The initial comparison of population-wise responses in response to accurate thalamic input for single whisker deflections was essential to demonstrating that the model was closer to in vivo, and we were unaware of similar data for non-barrel somatosensory regions. Moreover, our optogenetic & lesion study demonstrated the capacity to compare and extend studies of canonical cortical processing in the whisker system.

However, we cannot always predict the effect caused by mixing data sources. An interesting example of this is described in the companion paper (Reimann et al., 2024), stating that certain rules of inhibitory connectivity found in a mouse electron-microscopic (EM) connectome (Schneider-Mizell et al., 2023) cannot be explained by an non-selective pruning axo-dendritic appositions. We demonstrated the utility of the model to explore the effect of any connectivity rule, using the example of a mouse EM-compatible “SM-connectome”. After rewiring, and refitting synaptic, and conductance injection parameters, we found that the purely perisomatic targeting by PV+ neurons and the increased inhibitory targeting by VIP+ neurons shifted how much individual layers were driven by intrinsic and extrinsic excitation. This could not be explored with current techniques in vivo and highlights a strength of in silico experimentation. On the other hand, this data source contradicts some of our rat nbS1 parameters, e.g., the number of synapses per connection from light microscopy (Markram et al., 2015) do not match with the ones obtained from mouse V1 EM. Thus we used the rat data sources for building our baseline model, but make both models available in a simulatable format (see Data Availability).